\chapter[Sermon 54. The Dragon Persecutes the Woman: She Is Defended and Preserved of the Lord. The Dragon Stands on the Sand, Etc.]{The Dragon Persecutes the Woman: She Is Defended and Preserved of the Lord. The Dragon Stands on the Sand, Etc.}
\label{sermon:054}
\markright{Sermon 54. The Dragon Persecutes the Woman: She Is Defended and ...}

And when the dragon saw that he was cast down to the earth, he persecuted the woman who bore the male child. And to the woman were given two wings of a great eagle, that she might fly into the wilderness, to her place where she is nourished for a time, times, and half a time, from the face of the serpent. And the dragon cast water out of his mouth, as a river, after the woman, that he might carry her away with the flood. And the earth helped the woman, and the earth opened her mouth and swallowed up the river which the dragon cast out of his mouth. And the dragon was angry with the woman, and went and made war with the remnant of her offspring, who keep the commandments of God and hold the testimony of Jesus Christ, and he stood on the sand of the sea.

Having begun to speak of the persecution of the Dragon and the flight of the church, and having deferred it briefly to declare Christ’s victory, he now resumes and completes that discussion, vividly and most clearly describing the battle, and repeatedly emphasizing the help of God, which is given to the church through God’s grace.

When Satan, therefore, saw himself overcome by Christ and completely cast out, he began to rage against the church redeemed with the blood of God’s Son and vexed her with grievous persecution. For immediately after Christ’s ascension, a great persecution was stirred up against the Apostles and the Apostolic church. The Apostles, put in prisons, were grievously rebuked with words, and also scourged with rods and whipped. Stephen was stoned, James beheaded with the sword, finally, by the means of Paul (who then was called Saul), innumerable were cast into prison and put to cruel torments.

On the contrary, he recounts the present help of God, which he expresses through a figurative kind of speech, following the manner of the vision for greater effectiveness. For he says that to the woman, meaning the church, were given two wings of a great eagle, by whose aid she fled into the wilderness, where she concealed herself for a time, safe from the dragon’s sight. And here is signified that a great power is granted to the church to flee and escape the fury of Christ’s enemies, and to proclaim the gospel among the Gentiles.

Whereof you may read in the eleventh chapter of the Acts of the Apostles. And he does not mention the eagle’s wings without cause, and that of a great eagle. For Moses in Deuteronomy makes mention under this figure of the defense and aid of Almighty God, just as the eagle flies over her young, and stretches abroad her wings, and carries them on her shoulders, so the Lord has also kept and advanced them.

\index{Judgment, Last}\index{Daniel (Book of)}Furthermore, he says there is a place given to the church in the wilderness, namely, provided by Christ, which turns to him whom he will, and prepares for himself a spouse. And he nourishes the church among the Gentiles with his evangelical word, as he nourished his people in old time in the desert with manna. And the time of the church he does not prescribe. For he uses again a kind of speech as it were a riddle, borrowed out of Daniel, which God employs when he wishes the time to remain unknown to us; which, since we know that good and just things consist in him, we should not curiously inquire after. Of this I have spoken before. Doubtless it is plain that the church among the Gentiles shall continue and remain until the last judgment. But the day of judgment no man can define.

Again he declares, with what fury the Devil inflamed, shall make a new and continual war against the church. When he saw the church among the gentiles to be daily increased and established, he vomited after the woman water: and that we might know the figure, he adds, as it were a river. For he signifies that the Devil has unleashed a sea of evils into the church, sects I mean, dissensions, tumults, seditions, persecutions, wherewith the whole world has overflowed. Verily he raised up everywhere all magistrates and priests against the Apostles and Apostolic doctrine. Read the Acts of the Apostles chapters 13, 14, 15, and the chapter following. Neither is it rare in the Psalms, by waters, floods, and rivers, to understand all kinds of afflictions. And to this end he raised up those great evils, and unleashed them upon the godly, [illegible], that he might cause the church to be carried away by the flood: that is, that he might take away the godly and the doctrine of piety. And this is the continual endeavor of Satan; hereunto he applies all his consultations and doings. So in the empires of Nero and Domitian, he studied to wash away the church by the blood of saints, but yet in vain. For therefore I suppose it is said, he vomited a flood of evils after the woman, not upon the woman.

\index{John, Saint}For God never failed his afflicted church. It is a thing to marvel at, that the earth opened her mouth and swallowed the flood poured out from the serpent’s mouth. This earth drank up in old time and covered the blood of Abel. Here is signified that the godly, enduring persecution, have help from an unexpected source, as David is recorded to have been delivered by the help of the Philistines, expecting nothing less than to deliver David from the hands of King Saul. But while they intended another thing, they brought to pass that which seemed good to the Lord, who can turn the evil counsels of evil men to the profit of the godly. And we see many times in the Acts of the Apostles that the earth has swallowed a flood of evils; that is to say, that worldly men, doing something else in the meantime, have procured peace for the church. So did the town clerk or recorder of Ephesus pacify the multitude of Ephesians, who were all in an uproar and worse than mad. Lysias, the captain of the guard, takes Paul away from the bloody hands of the Jews; so did the centurion defend Paul, that he should not be slain by the soldiers in the shipwreck. The civil wars that began immediately after the death of Nero gave peace to the church until the reign of Domitian. But the old serpent, who can never rest, attempts new wars. For now, being mad with rage at the church, he goes to make war against the remnant of the woman’s seed, which is truly to be borne until the judgment of the church by the word. And so he makes way to the Roman persecutions, which followed immediately after the time of Saint John, in the Roman Empire, and the Antichristian persecutions raised after the Empire was subverted. This will be spoken of in chapter 13 and following verses.

Nevertheless, from this it chiefly appears what Saint John understands by the woman, namely that which engenders the seed of God. The church is called both Mother and daughter. The daughter, because she is engendered by the word preached in the church; the Mother, for that by the word, she brings forth spiritual children to Christ. For the seed of God, and the seed of the woman, are all those who keep the commandments of God and have the testimony of Jesus Christ. They keep the commandments of God, which esteem God’s law and frame all parts of their life according to the same. They do not keep God’s commandments, which set nothing by the law or word of God, nor frame their life after the same. Of this matter, there is spoken at large in the fourteenth [?chapter]. The testimony of Jesus Christ is nothing else but the gospel of Jesus Christ, preaching to us the free remission of sins. They have this, which possess it by faith.

And where he says, that the dragon stood on the sea shore, it is a preparation for what follows: for presently he says how the beast, the principal instrument of the dragon, came out of the sea by the devil’s means. And it is a comfort that the dragon is said to stand on the sand, and not on the rock. For it signifies that the fury of Satan shall not long endure against the church, and that the kingdom of the devil shall be shaken and fall to decay, whose foundations are laid upon the sand.
