\chapter[Sermon 76. Again This Vision Is More Fully Declared, and the Punishment of the Beast Is Shown.]{Again This Vision Is More Fully Declared, and the Punishment of the Beast Is Shown.}
\label{sermon:076}
\markright{Sermon 76. Again This Vision Is More Fully Declared, and the ...}

And he said to me: The waters which you saw, where the prostitute sits, are peoples and nations, and tribes, and languages. And the horns which you saw on the beast are those who will hate the prostitute and will make her desolate and naked, and will devour her flesh, and burn her with fire. For God has put in their hearts to fulfill his will, and to do with one accord, to give her kingdom to the beast, until the words of God are fulfilled. And the woman which you saw is that great city which reigns over the kings of the earth.

The angel sent by the Lord Christ reveals to John and the entire world the mystery of the beast, but especially her judgment or punishment for her heinous crimes. He will pursue this also in the following chapter.

And now he explains the meaning of the waters, over which the prostitute holds sway, namely the Roman power. Waters signify kingdoms dispersed throughout the world. He explains this in his own prophetic manner, as was noted before, by three words. For in naming people, folk, nations, and tongues, he encompasses as it were countless nations, distinct with diverse languages and customs. But where nothing is more changeable or unstable than waters, and when they are stirred up, they become more furious and outrageous: the common people are rightly compared to waters, which are also called changeable or unstable because of their fickleness, and furious and mad because of their rage.

Therefore, not without cause, all wise men have severely condemned seditions, which we are accustomed to call tumults or uprisings. For through such disturbances, many wicked individuals gather, and opportunity is given them to act on their desires and to inflict harm. But if many nations were subject to the Roman Empire and nevertheless strayed from the true faith, what will it avail to count up many kingdoms that might agree on a religion? As though the sincerity and truth of religion depended on a multitude of people agreeing in the same.

\index{Rome}Now follows God’s judgment against bloody Rome, which is the chiefest thing in this vision: the sum of it is, Rome shall be rent in pieces and burnt with fire, as we heard also in chapter 13. Just as she has done to others, so shall be done unto her. And these things are to be explained first of old Rome, and afterward of new; and in the same, the words must first be considered, than a conference of histories must be had, out of which the truth of the prophecy may appear.

Ten horns signify kings who arose from the division of the Roman Empire, such as the kings of the West Goths, East Goths, Germans, French, Lombards, Huns, Vandals, and so on. These nations served the Romans and received payment from them, favored them, and to their own detriment carried out their business, no differently than an earnest lover serves a harlot, from whom he cannot be withdrawn, whom he loves most fervently, but at last, perceiving her deceitful behavior, he begins to hate her most intensely. So these nations and others began to persecute the name of the Romans, desiring that no monuments or traces of them remain anywhere. All the provinces of the Romans were filled with Roman statues, images, pillars, inscriptions, and writings; but in those places, especially in Germany and its borders, how many of such great abundance remain? The cities where the Romans had their garrisons are utterly destroyed, so that scarcely any traces of them appear today.

And just as an honest man, having a prostitute as his wife, a shameless strumpet, does not only hate but also forsakes her, troubles her, and strips her bare, having plucked from her all her wifely apparel and ornaments (for so God in his prophets threatens to do to his people for their unfaithfulness), so nations revolted from the Roman Empire, destroyed and impoverished it, spoiling the riches thereof, which the Romans had heaped together by the robbery of all nations; they spoiled everywhere also the Roman provinces. And where it is said that those kings shall devour the flesh of the beast, it is to be understood in the manner of speaking. For so we are wont to say, when we signify extreme cruelty and malice without mercy; therefore, just as Rome has been most cruel towards all nations, even so shall all nations most cruelly tear her, and finally shall burn her with fire.

\index{John, Saint}Now let us compare these things with history, and see how they were fulfilled in old Rome, and may be yet fulfilled in the new. And first we will speak of old Rome, and after of new. And truly, old Rome grew many years, and practiced robberies throughout the whole world, and destroyed the saints of the Most High; wherefore it was worthy that the punishment thereof should extend and endure many years, and so as it were by degrees to descend to the last burning and destruction thereof. There are gathered the years of her punishments about 136, in which she, being impenitent, was vexed and tormented with continual calamities, slaughters, and vexations. And hereof I compiled an abridgement in the 57th Sermon of this work, the 13th Chapter. And here I will repeat a few things, and will rehearse certain other things more plainly and at large. As the Lord in punishing the Ninevites and Jerusalemites, declared his long suffering and clemency, and also his straight justice; right so in proceeding slowly to destroy Rome, he left them mercifully space to repent. But since they refused to do so, he wasted and destroyed them terribly as impenitent. He gave therefore to Rome excellent good rulers, Constantine, Iovian, Valentinian, Gratian, Theodosius, and so forth, by whose diligent labor and godliness he disclosed the furies and raging idolatry of the heathens, and also restored and established the true religion. But as in the time of Josiah the old, ingrained error and abominable idolatry could not be rooted out of their hearts, but that the greater part would rather keep the abominations of the Amorrites; so the Romans both in the city and in the provinces eagerly aspired to the restoration of old idolatry. Therefore, just as he tamed at length with grievous wars the invincible ungodliness of the Jews, and destroyed the city of Jerusalem, so by the war of the Goths and Vandals, and invasions of barbarian nations (as the histories term them), he destroyed proud and wicked Rome, with her provinces, and finally consumed the city with the sword and fire of the Goths. The very name of the enemy cried out that the vengeance was not executed by men, but by God himself. For the Germanic word for Goths signifies the people of God, or gods’ people. For God in Old High German is called \textit{Got}; from that comes \textit{Gotthes}, \textit{Die Gotther}, the people of God. Therefore God, and not man, did chasten, torment, and at last destroy Rome. Which thing John the Apostle speaks most expressly at this present.

\index{Jerome, Saint}First, during the rule of Honorius and Arcadius, the Visigoths, through Alaric, besieged, attacked, captured, and plundered the city. Jerome to Principia greatly laments this misfortune of Rome in the Epitaph of Marcella; but Orosius, as I also mentioned in sermon 57, in my judgment more rightly commends the just judgment of God in the affliction of Rome. It is clear that Rome was then, because of the enormity of her sins, chastised with mercy, but where the Romans would not acknowledge the hand of the one who struck her, it came to pass that Alaric, being dead, the victorious army, now led by Adolphe, returned from Lucania and plundered the riches of Rome even more greedily than before.

\index{Papacy}\index{Repentance}From that time, Rome was again granted a period of repentance, about forty-two years. In the meantime, through destruction and devastation brought by the Huns in their provinces, and that great and wondrous event, they were admonished to be wise. What will they say who observe that Attila himself, with his Huns, now invades Italy itself, and now hangs over the neck of Rome? Then there occurred an event that could have turned the Romans to the service of the true God, had there remained in them even a spark of gratitude. For the minister of the church of Rome, Bishop Leo (the ambitious pride of Popes was not yet known), a preacher of the Christian faith and a steward of Christ's mysteries, making supplication to Attila, obtained peace for Rome, and by a manifest oration turned away the bloody enemy from the necks of the Romans. This was an exceedingly great benefit, which God by his servant would have shown to the Romans, if they would yet cease to hate the religion of Christ and to slander Christ, as though he poured out evils into the world, and that no good nor quietness came from the preaching of the gospel. For even now (not to speak of others innumerable), he has implored upon them a benefit inestimable, and that by the preacher of the gospel. This was done in the year of our Lord 454.

However, while the Romans proceeded according to their accustomed manner, and now also Valentinian, a prince not evil, was murdered, and by a riot many dishonorable things were done, nor did any sign of thankfulness toward Christ appear, or sign of true conversion. Through the influence of one woman, Eudoxia, the wife of Valentinian, who herself also suffered many dishonorable things in that riot, it came to pass that Genseric, king of the Vandals, sailed out of Africa with three hundred thousand men to Rome, and captured it, and for the space of forty days, he gathered up the treasures brought there from all parts of the inhabited world. Then the intercession of Leo could do nothing, except that the Vandals abstained from killing and burning: which was also a benefit of God not to be scorned. The first king of the West Goths who broke into Rome was called Alaric; others call him Atalaric; but this king of the Vandals is named Genseric, and so Rome, a harlot, is made desolate and naked, plundered, I say, which, having been enriched with the spoils of all nations, was hitherto proud. However, it was not now altogether destroyed and burned: which was no small benefit, which Christ again showed to Rome for an amendment.

And yet moreover, about twenty years have passed, during which time, nevertheless, as with the ten tribes of Israel before the destruction of Samaria, continual murders were committed while princes reigned in Rome. Yet, for all that, none of these rulers escaped being either slain, murdered, or expelled. Augustulus was the last. For just as Augustulus succeeded Julius and initiated the Roman monarchy, so Augustulus brought it to an end.

For the Roman legions being extinguished, and the imperial title taken by the Germanic people, Odacer, (who took his name from destroying lands, Oedacher, and was called as if a destroyer,) captured Rome, and reigned there as king for about fifteen years. Yet he was expelled again, and slain (at the instigation of Zenon, Emperor of Constantinople) by Theodoric, Prince of the Ostrogoths. And the Ostrogoths reigned at Rome for about fifty years, until the Emperor Justinian sent Belisarius into Italy with a Greek army to recover it. The Ostrogoths, being aided by a force of Germans sent to them by Theodeper, king of France, valiantly resisted. They warred in Italy for eighteen years, with fluctuating fortune.

\index{Augustine, Saint}At the last, Totila overcame Baldeuille. He seized and burned the city of Rome, but not suddenly. For he gave a time to deliberate. But where he could not so prevail, he destroyed Rome, and as John has prophesied, burned her with fire. All histories mention this destruction. John Augustine in the third book of Chronicles has written of the same matter: Totila besieged Rome and took it, on the sixteenth day of January (the seventeenth of December) in the year of Christian salvation 548. Totila gave all the goods to the soldiers, but he commanded by proclamation that their bodies should be free. From there, he sent ambassadors with his wishes to New Rome (Constantinople) to Justinian. He demanded of the emperor Italy and the league as it had been under the emperors Anastasius and Theoderic, king. Which if he could not obtain, Totila threatened that he would raze the city, which he could not keep, and abolish the Roman name. Justinian answered that Belisarius was in Italy, to whom he had committed Italian matters. Totila, therefore, where the emperor would not grant his requests, determined to raze the city of Rome. He leveled the greatest part of the walls in most places and set the Capitol building on fire. He commanded all citizens with their wives and children to depart from the city. The common people of Rome were dispersed in the towns of Campania. The senators and nobility Totila kept with him as hostages. Then fire was put into every house. Thus, Rome being fired in all places, Totila left it vacant. For thirteen days, the fire burned freely. The city of Rome was forty days in that solitude, that there was neither man nor woman in it. After destroying the city, he moved his camp toward Lucania and Calabria. Belisarius came to the city left vacant, and sooner than one would have thought, fortified a part of the city with a ditch, wall, rampart, and wooden towers, for all could not be restored. Totila was with him, but repulsed, departed to Tibur. Belisarius was sent for into Greece by the emperor. Totila besieged Rome and took it. So in one year, Rome, the head of the world, the lady of all nations, was taken three times, thus writes Augustine. Leonard Aretine, writing of the Italian war against the Goths, in the end of the second book, says: After this, Totila departing from Rome with his whole army, left it utterly desolate and vacant, and so on. Who will say now that John has not in few words comprehended the destruction of old Rome, which the histories afterward have plentifully described: and finally, how, in the same manner as it was prophesied, it has followed the prophecy, after 451 years? And that so evidently, that you would think presently to behold Rome both falling and burning.

\index{Antichrist}\index{Turks}And just as in the story of the gospel the Lord intermixes a prophecy of the destruction of Jerusalem and of the end of the world, so that every man might, seeing the city of Jerusalem perish as he had prophesied, with no stone remaining upon another, gather with like truth and certainty that this world shall fall: so may we, seeing the once great city of Rome fallen and so great an empire, which was thought should have lasted forever, brought to naught, gather also that new Rome with her shadow or image of the empire shall as sure as day fall and be brought to naught. And first indeed the Saracens and Turks, who ruled and yet reign in the provinces subject to the Roman Empire, as in Asia, Greece, Egypt, Africa, Slavonia, and Hungary, and therefore are rightly accounted among the ten horns, do hate worse than dog or snake both Popery itself and Rome, and all that imagery of empire. Yea, and histories also testify that they have often made invasions and spoiled Rome itself. What is done at this day, experience itself teaches. But whether the Turk or the Christian princes themselves, converted to Christ by the gospel, shall spoil this new Rome, destroy it utterly and burn it with fire, the Lord knows, who seems here to intimate some such thing hereof. This is certain that Christ alone with his hand shall bring down Antichrist and abolish him with his coming. Certain it is that the Earth and all the works that are therein shall be broken. For thus is the apostolic doctrine: and that all these things shall be in the end of the world. Read Paul, 2 to the Thessalonians, chapter 2, and Peter, 2 Epistle, chapter 3. Moreover, there arise in sundry kingdoms of the world learned men, who, once bounden to the See of Rome, have defended her and her stinking idols; but after being converted to Christ, begin to hate both Rome and the Roman church, which also they assail and burn with the fire of God’s word. Therefore all the glory, dignity, and wealth of the Pope and Popery has perished and perishes daily in the godly. All that are godly and wise hate Rome and Roman wares. All cry out that this Sodom is worthy to be burned with fire falling from heaven. Neither is there any doubt but that a grievous vengeance is prepared against her.

And briefly is shown a reason wherefore the kings should rage so cruelly against the beast, and why these things are done in such sort and manner as we have heard. For God says he has given into the hearts of the kings, that is, that they should work his will and should do with one mind and consent. For where some refer this to the beast, that seems far off and strange. It is referred rather to the next, to God I mean, who put into the hearts of the kings to do his will, I say of God. For it is the mind and will of God that the beast should perish, that punishment be taken from her for shedding of innocent blood. The same God will procure that kings shall not be at discord, but at concord, that being of one mind and accord, they may execute God’s judgment. So we read in the Prophets that God put into the hearts of kings, Shalmaneser, Sennacherib, Nebuchadnezzar, Cyrus and others, that they should do as they are said to have done, namely in punishing the wicked and defending the godly. And there is also mention made in Histories how Alaric, king of the West Goths, was indeed dissuaded by a servant of God that he should not make such haste to destroy Rome, but that he answered, “There is one that continually troubles me and says, ‘Go, destroy Rome.’” And he that put that mind and will into the heart of Alaric, Adolphe, Genseric, Odacer, Theodoric, and Totila, the same if he will, and when he will, and into what princes he will, shall put, that they also shall do their duty against this new city and church of Rome.

\index{Daniel (Book of)}The angel adds that God has also put into the hearts of kings, so that they should give their kingdoms to the beast, until the words of God are fulfilled. The interpreters explain and say that God has permitted them to conceive this counsel in their minds, to deliver the kingdom to the beast. But I believe it is clearer if we simply confess that God is the author of no sin, and that men sin not as compelled by any fatal necessity, but through their own fault and vice. Therefore, God would, as he has also expressed and taught in his word, that kings should deliver their kingdoms to Christ, the supreme King. But since it did not please them, and they preferred, for various worldly and fleshly reasons, to deliver their kingdoms to the Pope, and submit themselves to the so-called Apostolic See, God, in his just judgment, has forsaken them and given them over (as Saint Paul wrote to the Romans) into a reprobate mind, to do those things which God does not allow. And so the words of God, prophesied by the Prophets and Apostles, are fulfilled in this way. Surely these are the words of God, and not of men, which are read concerning this matter in Daniel and throughout this book of Revelation.

Finally, the angel explains what is signified by the woman sitting on the beast: namely, that great city of Rome, the head and leading power of the world, and the Roman church, Papacy, and authority, extending herself and her dominion over the kings of the earth. Of this we have already spoken sufficiently. To God be glory.
