\chapter[Sermon 87. Of the Bright Verity of the Gospel, Which by the Ministry of the Apostles Was Spread Abroad Throughout the Whole World, and by a Thousand Years.]{Of the Bright Verity of the Gospel, Which by the Ministry of the Apostles Was Spread Abroad Throughout the Whole World, and by a Thousand Years.}
\label{sermon:087}
\markright{Sermon 87. Of the Bright Verity of the Gospel, Which by the ...}

And I saw an angel come down from heaven, having the key of the bottomless pit, and a great chain in his hand. And he seized the dragon, that old serpent (who is the devil and Satan), and bound him for a thousand years, and cast him into the bottomless pit. And he bound him and set a seal on him, so that he should deceive the nations no more until the thousand years were fulfilled. And after that, he must be released for a little season.

\index{John, Saint}Saying that the greatest aspects of religion and true godliness consist in the true knowledge and understanding of the final judgment, as I often advise: Saint John diligently discusses the treatise on the final judgment to our great benefit. And following his customary method, to ensure that all that he proposes is clear, he not only declares the matter in words but also presents it through visions as if they are events to be seen with the eyes—and this is for the faithful. For to the unfaithful, all these things, although most godly and divine, seem mere trifles and fables. But the wisdom of God will laugh at them also when she sees her time, as she threatens in the Proverbs of Solomon. And he also resolves certain questions that are commonly raised about this matter.

\index{Turks}And he said, how the beast with the false prophet and all his adherents should be cast at the last day into hell; but where the first part is neither Antichristian, nor yet Christian, but rather of their own sense and judgment, to be a rule and law to themselves—such as are the Nestorians, Jacobites, Georgians, and so forth—or those who are still heathens or gentiles, moreover Jews and Turks—some man might marvel and demand, what shall be done with them, or what will come of them? John answers: “And the remnant were slain with the sword of him that sat on the horse,” and so forth. Again, where a godly man might wonder how they should be condemned, who, being born among Turks, heretics, Jews, and gentiles, never heard the Christian truth? John anticipates this imagination and, at the beginning of chapter 20, shows with what majesty, clarity, and evidence the truth of Christ’s gospel was notified to the world; how also all force and power was taken away from the devil, and that for the space of a thousand years, wherein the preaching of the gospel thundered continually, so that those who have not received the gospel of Christ are utterly inexcusable. For the preaching of the gospel was not obscure, but most clear and manifest, nor short and contracted, but published for the space of a thousand years; it was not received by a few little ones, but by all people and nations under the sun. Therefore, it is a gross ignorance of the Turks, heretics, Jews, and gentiles. For although in times past the truth seemed to have been notably known, now it is not so; yet it is certain that the majesty of the gospel has been so great in the world that there is now mention of it with all men, and by their own malice they hide their eyes, which understand nothing of Christ. Therefore, that saying of the Apostle is even now of force: “If our gospel be hid, in those that perish it is hid, unto whom the God of this world hath blinded the minds of them, which believe not, that the light of the gospel should not shine unto them,” 2 Corinthians 4. Whereupon we now gather that none of those who are condemned in the world are condemned without deserving. Which thing the Apostle Paul also touched upon in the Epistle to the Romans in chapters 1 and 2. Here, therefore, is a profitable and necessary place treated, of the famous preaching of the gospel throughout the world, the course of which endured a thousand years.

And this treatise proceeds in this order. First, the angel is described, after which his work or effect is declared. And last is the sealing of the time. Concerning the description of the angel, first he is named an angel, and comes forth abroad; however, the entire apostolic state is understood here, in which Saint Paul, the doctor of the Gentiles, shines exceedingly. It is not surprising that the order of the apostles is signified by an angel, for an angel signifies a messenger, ambassador, or an apostle. And therefore, the prophet Malachi called John the Baptist, the herald of our Lord, an angel: “Behold, I send my angel before thee,” and so forth. Ministers of the church are often called angels in this book. But if the worthiness and nobility of the name pleases the ministers, let the angelic purity and excellent faith please them also. An ambassador does and says nothing except that which he has received in commission from him who sent him; so also let the ministers set forth nothing except that which they have received from the Lord, in the Scriptures.

Secondly, this excellent angel is said to come down from heaven, not that the bodies of the Apostles came from heaven, but because their vocation and office were given to them from heaven. For the Son of God, who came down from heaven, chose the Apostles and sent them forth into the world. This is declared in Matthew 10:20 and John 20:21, Mark 16:6, and Luke 24:48. And Paul says to the Galatians that he was called and ordained an apostle neither by men nor from men, but by God through Christ. Thus it appears how great is the authority of the Apostles. For they do not speak, but the Spirit of Christ and of the Father speaks in them. Therefore, he who despises their doctrine despises God the Father and the Son. They lie moreover, who say that the gospel is a new doctrine forged by clever men. Read the first and second epistles of Peter.

After this, the Angel is said to hold in his hand two excellent instruments, the key and chain. Let us see what is meant by them. Doubtless, by these two instruments John understands nothing else but the free, true, holy, and lively preaching of the gospel: by which it came to pass that both hell was locked from the faithful, and the devil was held and kept fast bound in chains, that he could not hurt the godly so much as he would, and seduce whom he listed. For so John will explain himself later.

And the keys of binding and loosing the apostles received from the Lord, in Matthew 16 and John 20. They open the bottomless pit and hell itself to the ungodly by preaching the gospel, when they show them their damnation in hell for their ungodliness. They shut up hell from the godly, while by preaching the gospel they open heaven and bring the faithful to celestial joys. I have spoken of the keys at length in another place. A chain is the sign of captivity. By the preaching of God’s word, the Devil is taken and bound, which is why common painters have depicted the Devil bound with chains to certain notable preachers.

\index{Rome}The things that follow this key, and the significance of the chain, are better explained, as is the effect of the apostolic preaching. For he adds: “And he seized the dragon,” referring back to chapter 12, where everything is explained. You may also look there for the same information. And the angel bound Satan, which is the purpose of the chain: to prevent him from stirring up, invading, and destroying the faithful. Furthermore, there follows something even more forceful: “And he cast him into the bottomless pit,” that is, he hurled him headlong into the depths of hell so that he could not murder the faithful. Another, even more serious thing follows: “And he shut him up so that he could not come out again.” This is the purpose and effect of the key. Furthermore, he placed a seal upon him. Letters, prisoners, and graves are typically sealed for security and credibility, to prevent tampering and to ensure they remain shut, sealed, and safe. All these things signify a complete and perfect victory that we have obtained through Christ, by the word of the gospel preached to us and communicated through faith. For our sake, he overcame, defeated, bound, imprisoned, and sealed the enemy, so that we might be safe and secure from him. Finally, it follows—this may explain all the preceding points—that he should deceive the people no more, by the same means he had previously used to seduce them before Christ’s victory and before the gospel was preached throughout the world. For then, everything was full of ungodliness and error. Temples of God or idols were everywhere, idols were worshipped, and they gave oracles. Altars smoked with the blood of men and beasts. All wickedness reigned. Arts, magic, witchcraft, and parricide were practiced without punishment. It is difficult to express, even in a lengthy speech, how shamefully Satan had deceived the world, how assuredly he reigned, and with what complexity he had bound humanity to himself like a slave. Let anyone who wishes examine Greece, Italy, and Asia; Corinth in Greece, Rome in Italy, and Ephesus, the leading city of Asia. They will find abominable abominations and say that the devil had reigned completely there, constantly bewitching the simple and wretched with new deceptions. But after Paul alone (I will say nothing now of the other apostles) came to Corinth, Ephesus, and Rome, and preached Christ, who cannot see that John here truly saw the devil bound and securely shut up? I am brief in this matter, which is abundant, because I think I have done enough if I have shown only a few stepping stones, by which one may come to a much more thorough consideration of these things. Relevant are the divine words of Paul, which are read in Acts 26, spoken before King Agrippa, the princes of Syria, and Festus, the proconsul. “For this purpose I have appeared to you,” the Lord says to Paul, “that I might appoint you a minister and a witness both of the things which you have seen, and of the things which I shall appear to you after this, delivering you from the people and nations to whom I now send you, that you may open their eyes, that they may turn from darkness to light, and from the power of Satan to God, that they may receive forgiveness of sins.” In 1 Colossians, as well as in other places, Paul shows that Christ has overcome Satan, and that the same Christ has redeemed us and brought us out of the kingdom of darkness into the kingdom and light of the Son of God. Therefore, when the apostles and ministers are said here to bind and shut up Satan, it is to be understood through the means of their ministry. Each person can also judge whether he has profited from the doctrine of the gospel that he has heard for a long time in the church. For if you yourself are still bound by the devil’s chain, you have not yet heard the gospel as it pertains. But if you feel that the devil is bound by the chain, and that you rule the devil, and the devil does not rule you, then things are going well. Pray to God: “Lord, confirm and increase what you have worked in us.”

\index{Eusebius of Caesarea}\index{Papias}\index{Millenarianism}And concerning the time of this most brilliant truth of the gospel, it is said how it shall endure in the world a thousand years. For he says expressly: he bound him for a thousand years. And again: that he should no more deceive the people, till the thousand years were fulfilled. I know that the opinions of the expositors, concerning these thousand years, are diverse. I know how the heresy of the Chiliasts or Millenaries, by Papias, the author of this view, as Eusebius recounts in the third book of the \textit{Ecclesiastical History}, was derived from this. I will not here linger to refute the opinions of others, which would also be overly long and tedious, and not of such great profit. I will only express my own view to be weighed by godly readers, and then I will leave it free for every man to follow that thing which he shall think most agreeable to the truth and profitable for the faithful. And I understand plainly and simply that Saint John speaks of a thousand years, which ran on by continual course from the time of Christ, until the last corrupting of the Evangelical preaching and church of Christ.

\index{Papacy}I am not particularly concerned with precisely determining the length of these thousand years. I simply set the beginning of the reckoning to the time of the open preaching of the Gospel, when the word began to be received by the Gentiles. Therefore, I suppose that there may be three periods appointed, which nevertheless will all fall within the same reckoning, differing little or nothing among themselves, or having only a small diversity, not exceeding half a year more or less. You may, if you wish, begin the calculation of the thousand years from the thirty-fourth year of Christ’s birth, when Christ also ascended into heaven, and Paul, being called to the ministry, began to restrain Satan by drawing the Gentiles into the fellowship of God’s people through the preaching of God’s word. And you will arrive at the year of our Lord 1034, and the papacy of Pope Benedict the ninth, who, after unlawfully seizing the chair of Saint Peter, as they call it, practiced magic and entered into league with the devil; of whom he was also carried away, as he had sold his bishopric before Pope Gregory the sixth. Read the story of Cardinal Benon, as mentioned previously in chapter 13, and let other stories be read. It is certain that the Devil occupied the apostolic seat at those times, as they term it. Read the stories from Sylvester the second and onward.

You will then say that about that time the Devil broke loose again and seduced the people, especially through Popes. Or begin the calculation of the thousand years from the time when Paul, being bound for the Gospel at Rome, testified that the Gospel had been preached throughout the world. That was about the year of our Lord 60. From that point, reckoning a thousand years, you will arrive at the year of our Lord 1060, when Nicholas the second was Pope, under whom it is written that the truth was diversely tempted and corrupted, and that Gregory the seventh also, by his schemes and endeavors, troubled the whole world. Or begin the calculation from the destruction of Jerusalem, when the Jews cast out the Gentiles, and a large number of Gentiles entered and were received into the place of the rejected Jews, which was the year of our Lord 73, and you will arrive at the year of our Lord 1073, even to Pope Gregory the seventh, in which time not a few historians write that the Devil himself reigned. Doubtless, no man has harmed godliness or more boldly advanced impiety than this Gregory, otherwise called Hildebrand. I have spoken of him before in chapter 13. There I also admonished you that Cardinal Benon calculated those thousand years from the birth of the Lord and concluded them in Sylvester the second.

It is evident, therefore, that the Gospel has held a notable place in the world and has not been extinguished for the space of a thousand years, that is to say, from the time of the apostles until the year of our Lord was reckoned 1073 or thereabouts. What was done at that time and afterward, we shall hear when we come to that saying: “And when the thousand years shall be fulfilled, \&c.”

\index{Antichrist}\index{Jerome, Saint}\index{Mahomet}\index{Justification}Some man will say, “I cannot see that the preaching of the Gospel has continued in the world for so long a time, namely a thousand years, for it appears by histories that the doctrine of merits, satisfactions, and justification by works, immediately after the time of the Apostles, laid their first foundations. We know that the intercessions of saints and the worship of relics were defended by Jerome, who departed this world in the year of our Lord’s incarnation 422. We know that the Bishop of Rome immediately after the death of Gregory the first, claimed to be head and catholic pastor of the universal church. We know that about the same time, namely about the year of our Lord’s incarnation 630, Muhammad seduced a great part of the world. We know that shortly after, arose that detestable contention about the having of images in the churches of Christians. We have heard that John has assigned years to Antichrist, 666. Finally, it is manifest that the Devil has by murder, parricide, and all kinds of mischief reigned in the children of unbelief. Wherefore you say, “I see not how the Devil has been bounden a thousand years and locked in chains.” I answer that the things which are alleged hitherto are true; yet nevertheless, to be true and so to remain always, which John by the revelation of Jesus Christ has affirmed, that the Devil should be shut up for a thousand years and remain bounden until a thousand years were at an end. And the same we declare on this wise. The Lord said in the Gospel, “Now is the judgment of the world; now shall the Prince of this world be cast out.” And whereas it is not lawful to doubt of the verity of Christ's words, yet nevertheless, he is not to be understood as being so cast out, but that he has been of great force in the world and has been called by the Apostles themselves, the Prince of this world. How then is he said to be cast out, and to tempt the godly, to reign, and to be cast out of his kingdom?

\index{Augustine, Saint}He is cast out of the church and of the faithful, not that he does not come again, for always he returns and seeks to pluck back the redeemed (but that he possesses no more the full dominion. For Christ now lives and reigns in the church and saints. These, as Saint Augustine says, he assails from without). He is cast out of his ancient possession, but he labors to recover his old habitation. And thus was Satan bound and shut up for a thousand years, as one who did not possess the faithful of Christ throughout the world, nor rule them at his pleasure and according to his malice, although he has tempted and vexed them. So was the Holy Spirit denied to be given, not that he was not in the world and in the Prophets, but because he was never so plentifully poured out upon all flesh as after the glorifying of our Lord Christ. In the same sense we say that death and sin are taken away from the faithful and trodden under foot. As Paul, who in the first chapter to the Colossians said that we are translated out of the kingdom of darkness into the kingdom of light, nevertheless says to the Corinthians that the god of this world has blinded the minds of the unfaithful. So John at this present says how the Devil is bound and sealed for the space of a thousand years, and the very same says afterward that the rest of the dead did not revive until the thousand years should be fulfilled: that is to say, in all those thousand years did not believe those who valued the beast more than they did Christ. And they, truly through their own fault and the instigation of the Devil, did not believe and perished. Therefore did Satan exercise his power in them. Which to the faithful in deed is bounden and tied fast, but to the unfaithful free and over familiar. Likewise, Hell is shut to the godly, to the wicked open. Wherefore also we confess in the creed, life everlasting and not death or everlasting damnation. For the faithful have no Hell, or there is no Hell prepared for them, but for the ungodly. For Christ has broken Hell, but for his faithful; to the unfaithful all things of Hell are yet most strong, and these have Hell.

Again, the Devil is said to be bound, shut up, and sealed: for since the redemption of Christ, his power has not been so great in the world as it was before. Therefore John explains himself, and says that he should deceive the people no more. What is this “more”? But that he shall not so seduce them from now on as he has done hitherto. Therefore, although in the meantime he shall deceive some, in those thousand years he has not reigned so fully, safely, and freely as he did before, and as it is permitted him after those thousand years to rage. Therefore, these things are spoken by comparison, and not absolutely. And the thing itself, or experience, teaches that they are not to be understood absolutely, and after the bare letter. Although therefore Satan has in these thousand years also blown his poison upon many, and has troubled the world, yet this is nothing in comparison of those things that have followed after the thousand years, even until this day, and shall follow hereafter unto the world’s end. In old time also he reigned fully among the Gentiles through idolatry. But a thousand years fell down their temples and idols, with all other instruments of ungodliness.

We read truly, how there were in the time of the Apostles who affirmed that people are justified by the law and works. Whereupon sprang up the doctrine of satisfaction and merits. But the same doctrine was refuted by the Apostle Paul, above other Apostles. Saint Augustine also, and after him Bede, most consistently have defended the doctrine of grace and redemption by Christ. This doctrine remained secure for the space of a thousand whole years. But afterward, monks gaining the upper hand, the doctrine of satisfaction and human merits did prevail; whereupon the doctrine of Jesus Christ concerning the free remission of sins and the imputation of righteousness was utterly obscured. There grew up an opinion with certain of the saints, making intercession or praying in heaven for their worshippers. Relics began to be worshipped too soon. Nevertheless, those who were illumined clung fast to the only intercessor, Christ, and did not honor relics. But after those thousand fatal years, many attributed more to saints than to the very holy one of saints. We see what is done at this day. The writings of monks and friars testify how much the worship of creatures has increased within these four hundred years, or thereabouts. Who will deny that exceeding many have been deceived by heretics? But who can gather thereby that the Gospel has been utterly lost, and that Satan has reigned fully?

The Bishop of Rome has ascended into the top of Mount Zion, and will be called the head and general pastor of the whole Catholic Church. However, the East consistently resisted, and so did other parts of the world as well. At length, after a thousand years, he made his boast most impudently, that the fullness of power was given him, which he obtained by trickery and clamor, and afterward usurped it. Muhammad seduced many, yet nevertheless the patriarchal churches persisted, and the East honored Christ, likewise the South and North, so that the thousand years again had their light; neither has Satan in these raged so much as he has since those years were complete. Doubtless, since the Turks began to rule and reign, all matters of religion grew every day worse and worse. And the war into the Holy Land did very much hurt to religion, and gave great courage to the Saracens and ungodly people; whereof I shall speak afterwards. And images began to be set up in temples, and to be defended. But the histories testify that the same was done with great difficulty, and hardly could the use of them be obtained, all good men most constantly resisting. And when they were now admitted, yet were not the idolaters so stark mad as we see they are now, and have been for certain years past. Wherefore it is rightly said how after a thousand years, Satan should be loosed from his chains, which before also moved the unbelievers, yet finally rages more furiously.

John assigned to the Antichrist a certain number of years, namely 666. From this, we might understand the name of the Antichrist. But it does not follow that the Devil was then completely loosed, or the light of the Gospel utterly extinguished. For the Apostle, speaking of him in his time, says, “The mystery of iniquity now works.” Therefore, the Antichrist has his seeds, his beginning, his rising, his growth, and his increase. But after a thousand years, he began to operate most impudently and boldly, which he had also done before, but now he spews out his poison most venomously, oppressing kings and all who speak against him, even if only a little.

We know moreover that in these past thousand years, the Devil has reigned in many through murder, perjury, and innumerable and unspeakable evils. But if you consider what has been done since those thousand years, and what is done today, you will say that those ages of the thousand years were like golden and silver worlds; and now for these five hundred years they are like brass, iron, lead, and clay.

Lactantius, in the seventh book of \textit{Institutions}, chapter 15, toward the end of the world, says that the state of worldly matters must necessarily be altered, and iniquity prevailing, to incline to the worse, so that these our times, in which iniquity and mischief have grown to the highest degree, yet in comparison of that incurable evil, may be accounted fortunate and, in a manner, golden ages. For justice will wear so thin, ungodliness, covetousness, stubbornness, and lust will be so common, that if there are happily any good men then, they will be a prey to the wicked and everywhere vexed by the unrighteous. And evil men only will be wealthy, and the good tormented in all vexation and misery. All right will be confounded, and laws will perish. Then no man will have anything save that which is either evil gotten or evil kept. Gold and violence will have all. There will be no faith in men, no peace, no humanity, no shamefastness, no truth. And the remnant which are read there. By all of which our days now seem to be painted vividly.

But what does it mean that the Devil must be loosed for a little season? Does this “little season” last now five hundred years? We interpret this passage similarly to the saying in the gospel: “Unless those days were shortened, no flesh should be saved.” For it is clear from history that neither the Devil nor Antichrist has enjoyed his kingdom without disturbance. In every age, holy and learned men, illuminated and comforted by God, like Enoch and Elijah, have resisted the ungodly and maintained the true religion. Through them, consciences afflicted by Antichrist have received comfort, God, in his mercy, tempering matters so that the elect should not despair in great temptations, errors, and darkness.

Therefore both Satan and the Pope could enjoy these matters for only a short time. For immediately after the thousand years sprang up the Waldensians, who constantly opposed the Pope and his ungodliness. The Lord has raised up certain kings, who are images—namely, the German Emperors Frederick, Lewis of the house of Bavaria, and many others. The Popes themselves have been at disagreement, while many have been chosen, and each of them will be the vicar of Christ, thus tearing apart that entire ecclesiastical body of theirs with schisms. Against these preachers arise Wycliffe, Huss, Jerome of Prague, and diverse others, with earnest and vehement opposition. What is done at this day, and has been happening for these thirty years and more, against superstitions and idolatry, against the Pope and all his clergy, the Papists themselves cry out, and all parts of the world can testify. Therefore the Devil has been cast down for a little season. The Lord Jesus will soon tread him under our feet.
