\chapter[Sermon 74. The Same Matter Is Yet Still Treated of, and the Vision Is Expounded.]{The Same Matter Is Yet Still Treated of, and the Vision Is Expounded.}
\label{sermon:074}
\markright{Sermon 74. The Same Matter Is Yet Still Treated of, and the ...}

And the woman was arrayed in purple and scarlet, and adorned with gold, precious stones, and pearls; and she held a golden cup filled with the abominations and filthiness of her fornication. And on her forehead was written a name, a mystery: Great Babylon, the mother of fornication and of the abominations of the earth. And I saw that the woman was drunk with the blood of saints, and with the blood of those who bear witness to Jesus. And when I saw her, I marveled greatly. And the angel said to me, “Why do you marvel? I will explain to you the mystery of the woman and of the beast with which she is carried, which has seven heads and ten horns.” The beast that you saw was, is not, and will ascend out of the bottomless pit, and go into destruction; and those who dwell on the earth will wonder—those whose names were not written in the book of life from the beginning of the world—when they behold the beast that was, is not, and will be. And here is a mind that has wisdom.

\index{Papacy}\index{Rome}He describes exceedingly well and vividly, setting forth for all men to see that very woman previously named the great whore. Toward the end of the chapter, he depicts death itself, and says: “And the woman which thou sawest is the great city.” Indeed, she is great, as one who holds dominion over the kings of the Earth. He means therefore the very city of Rome, and even the papal and Roman church, and the pope himself with all his officials and chaplains, who constitute a great and powerful city, ruling over all kings and princes of the Earth. For who does not know that Rome and the prelates of the church rule even above magistrates and princes? Consider what is done in the courts of kings and princes, and by whose advice and counsel the princes of the Earth are chiefly governed. And in calling papistry a woman, he alludes to the fifth and seventh chapters of the Proverbs of Solomon, who also likens crafty and deceitful philosophy, and worldly wisdom to a beautiful woman full of various schemes and deceptions.

\index{Antichrist}\index{Daniel (Book of)}And now also he clearly and vividly depicts the apparel and exceedingly immoral behavior of this woman. She is furnished with no good qualities inwardly, whereby she might commend herself to the world and to her lovers; therefore she excels in outward adornment, while inwardly she is full of all abominations. For she is like to all the world, like the Pharisees and hypocrites, whom the Lord says are outwardly shining tombs of marble, but within are filled with all corruption and filthiness, and even stench. And by this is chiefly signified that the church of Rome and the kingdom of Antichrist presents itself altogether with worldly trappings, namely with gold, silver, precious stones, and all costly array. These things are found among the people of old, under the discipline of the law; they are also found in the images of the gentiles, who suppose that God is not rightly honored, but with the precious things of this world. But we know that the Levitical priesthood is abrogated with all outward adornment, and that now the church decks herself with virtues, hates, and abhors outward embellishment. Lactantius refutes at length the external adornment in religious worship, in his book \textit{Institutes on True Worship}. Moreover, all ancient writers show that God is not worshipped by Christians with gold and silver, but with faith, charity, and righteousness. What will they say who claim that Daniel in chapter 11 shows that Antichrist will worship God with gold, silver, and precious things? This is condemned and rejected.

And does it not appear here plainly that the Lord Jesus himself has set forth to us the Pope or Antichrist, painted as it were in a table? For he appears altogether such, and in such like apparel he shows himself to be seen of all men, as the whore of Babylon is decked at this present. And he challenges to himself this apparel by a certain right. For the Papists bring forth a false, feigned donation of Constantine, among other things, pronouncing thus, in Distinct. 96. We give and dismiss to blessed Sylvester and to all his successors the Lateran palace of our Empire; moreover, the diadem, to wit, the royal crown of our head (which the Pope calls a kingdom, and has made it triple) and also our purple robe and coat of crimson, and all our imperial array, \&c. I cannot here omit, but must needs write out a few things of Platina, the Pope’s secretary, \textit{De Pontificibus}. For describing the life of Clement V at the Pope’s coronation, he says that Philip, King of France, and Charles his brother, John, Duke of Britain, were present, who, overwhelmed by the falling of a wall, dies (see the divine and just judgment of God) with many others, whilst the pomp of the Coronation, as the manner is, was led through the city. King Philip also by the same ruin was sore hurt and lamed; the Pope, stricken from his horse, lost a ruby out of his mitre that cost six thousand ducats.

\index{Ezekiel (Book of)}This harlot moreover drinks to all nations from Circe’s cup, which the Lord calls of Gold. And it signifies doctrine. For to give drink is to teach, as in Ezekiel. Gold betokens sincerity and purity of doctrine. Doubtless under pretense of sincerity and divine truth, Rome has easily persuaded all people to receive the doctrine of the Roman See. For the Pope has named himself Apostolic, and the church of Rome also Apostolic. And in the Canons has left written: so are all the laws of the See to be taken as though they were confirmed by the godly mouth of Saint Peter himself. Distinct. 19. Read chapters 20, 21, and 22. Distinct. Therefore the more simple sort of the world have supposed that they receive the very word and laws of God, when they received the decrees and doctrine of the church of Rome. But our Lord Jesus Christ declares here what has been, and what is even at this day their doctrine, and says: full of abominations and uncleanness of her filthy lust. And the scripture calls abomination, idolatry, as in Deuteronomy, chapter 7. Moreover, the false worship of God, superstition, and such things like. The uncleanness of whoredom in the Prophets is nothing else but perverse doctrine and perverse religion, not attributing all good things to God alone by his Son, but rather dividing the heart and applying them both to creatures and to wicked worships. But such is the doctrine and religion of the Roman See. Therefore her great sin is here recited, that she has with her evil and venomous doctrine seduced and infected all nations, and even now retains them in superstition and idolatry. A like place is in chapter 16 of Ezekiel. And I suppose that our Lord Jesus Christ used here now words very filthy, to the intent he might pluck away from the Roman decrees and Canons their authority and visibility, and that their filthiness might appear unto all men, and be known, and eschewed.

Furthermore, lest anyone should be ignorant of what the woman is, who is here portrayed for view, and that all might flee that great witch Circe, he writes her own name on her forehead, so that all men might read it, and that she might by no means be unknown. For he calls a mystery the understanding or signification of a secret; for by a figure of speech Rome is called Babylon. Whereof I have spoken before. And after the true signification of the word, Babel signifies confusion. And Rome has brought an exceeding great confusion into the church. For just as the primitive church of Rome set forth the gospel in the western lands, so now, the first simplicity and purity being extinguished, later bishops, regarding more ambition and covetousness than humility, liberality, and godliness, have brought into the whole world all manner of idolatry and superstition. Certainly, she is called by the Lord expressly the mother of the harlotries and abominations of the earth. For we may thank the church of Rome for all the corrupt doctrine and ungodliness that is in the church. She is the origin of idols, the Mass, and other abominations. Therefore, she is most worthy to be punished with most grievous torments. And this verily is the most worthy title of the Roman church. Others call her Apostolic, divine, chief, and of all most holy; the Lord Christ calls her Babylon, and that great harlot, and even the mother of the abominations and harlotries of the whole world, fighting against God and his Anointed. Therefore, let all the holy and obedient children of God the Father flee from her. I will not here make rehearsal of how she is also the mother of all abominations and harlotries, even after the flesh. For whilst the church of Rome has prohibited lawful marriages and, permitted by God to ecclesiastical persons, it has opened the gates to fornications, adulteries, harlotries, and lusts abominable. There need no words; the thing itself speaks.

To these, he adds a crime of all others the most grievous. And here also he employs an amplification. For he says that great Circe, the most venomous witch and sorcerer, is not sprinkled, or immersed, or wet, but drunk with the blood of saints, that is, of holy martyrs, who have borne witness to Jesus Christ by preaching the gospel and ascribing all things of salvation to Christ alone. But how many thousands, nay millions of martyrs, through the instigation and means of the Bishop and church of Rome, have been executed with most extreme and horrible punishments within these six or five hundred years, histories make mention. What has been done, and what abundance of human blood has been shed even within these thirty years, which our memory reaches, my heart grudges to recite. Great is this crime also, for which Babylon is plagued by God with just and most grievous torments. And aptly here is made mention of martyrs, that is, of the witnesses of Jesus. For those who confess the Evangelical doctrine of Christ to be the true and absolute doctrine, that Christ is the only head of the church, the only priest and bishop, mediator and sacrifice, and will not join with all, while at the same time that the doctrine of the church of Rome is also most perfect, to be held in like reverence with the doctrine of the gospel, that the Pope is head of the church militant, and on earth the true vicar of Christ and general pastor, and that the saints in heaven pray for us, and that the mass is a true and real sacrifice for the sins of the living and the dead—they are condemned as heretics and schismatics, with sword and fire to be rooted out of the earth.

And thus far extends the mystery of the vision, with terrifying sights presented before our eyes, wherein is vividly described both the ancient Roman Empire and, principally, the Papacy of Rome with its sins and crimes, heinous and full of enormity. Afterward shall follow an exposition of the vision, to which at the last shall be annexed the punishment to be taken of Antichrist and the entire Antichristian City. But if anyone applies all these things which are spoken of the woman to ancient Rome, I will not object. For there was also a religion at Rome, which consisted in gold and precious things. Ancient Rome had a cup of false wisdom, with which she made drunken and infected all nations. She was therefore the mother of abominations and whoredoms, from whom the provinces learned superstitions, and so forth. Nevertheless, these things particularly concern the Pope. Notwithstanding that ancient Rome also was drunken with the blood of Saints, and so forth.

John was greatly astonished when he saw the woman. Daniel also was astonished, so much so that his heart almost failed him, when he saw that Roman beast, as appears in the seventh chapter of Daniel. Godly people are also astonished today when they see so great things tolerated or permitted by God to the church of Rome against the pure and sincere. For the prelates of the church are fortunate, victorious, powerful, and in favor with all princes, and they bring to pass whatever they imagine or desire. Happy is he who is not offended by them. Read the seventy-third Psalm. How good God is to Israel, to those who are upright in heart, \&c.

The Angel, as chief minister, reveals the mystery to John and the whole Church: that is to say, he opens the secret and true meaning of the vision by parts most diligently. And he speaks in deed of the whole body of the beast, notwithstanding that the beast has certain things peculiar, and likewise the whore. Yet the Angel himself says, “I will show thee the mystery of the woman, and of the beast that bears her.” Nevertheless, this is also a dark speaking, wherein in the beginning of the exposition he says, “The beast which thou sawest was, and is not.” The Roman Empire was still present while Domitian ruled, but it was no more such as it had been before. For from the first monarch Julius, it was as it were by inheritance in the house of Caesars, until Nero. For in him the beast received a deadly wound, but it was healed, and diverse emperors reigned not of any one family. The Empire therefore had been in the power of one house before, but after Nero it was not so. Again, the Romans possessed the Empire after Nero. From Nerva, which is the seventh after Nero, the Empire was devolved to Trajan, under whom it was puissant and strong. Therefore it was and it was not. Whereof John himself will speak a little after. Moreover, the histories testify that the Empire of Rome was extinguished, and in its place sprang up another, which is also called the Roman Empire; whereof you may say most truly it was, and is not. For that old Roman Empire was the most ample and noblest Empire in the world, but this new, now erected by the Pope, is none such, but rather an image of the beast (as I said in the thirteenth chapter), a shadow and a dream. Therefore we doubt nothing but that in this vision is exhibited to us a type, both of the old and new Empire, but chiefly of Papistry.

And now he demonstrates no other origin of the beast but a hellish and devilish one. For he says how it shall arise from the bottomless pit, as was spoken of before. All empires, certainly, as Daniel shows in chapter 2, are of God. But if the rulers are corrupted, the origin is attributed to the Devil, not of the empire itself, but of the corruption. Moreover, it is added here what end that unfortunate empire is likely to have at the last: and it goes into perdition. For it is cut up by the roots on earth, and in the next world is condemned to everlasting punishment.

But just as he demonstrated in the thirteenth chapter, concerning those who should wonder at—that is, honor and worship—the beast, so here he repeats the same, not the elect children of God, but earthly men and those condemned, whose names are not written in the book of life. Of this we have spoken before. He adds an exclamation: and here is a mind that possesses wisdom. The Lord urges all hearers to the diligent consideration of these things, lest we be deceived and perish. Those are fools who marvel at the beast’s prosperity, victories, the Pope’s majesty, riches, and pleasures, and submit themselves to him. Truly wise are those who understand that the empire is taken away, and that under the guise of an empire lurks Antichrist, the child of perdition and man of sin, to be shunned by all the faithful. For these have turned to Christ, in whom they know they have all things pertaining to life and salvation, and to live in him. To him be praise and glory.
