\chapter[Sermon 58. Of Another Beast, Which Comes Up Out of the Earth: That Is to Say, of Antichrist.]{Of Another Beast, Which Comes Up Out of the Earth: That Is to Say, of Antichrist.}
\label{sermon:058}
\markright{Sermon 58. Of Another Beast, Which Comes Up Out of the Earth ...}

And I saw another beast rise up out of the earth, and he had two horns like a lamb, but he spoke as the dragon.

The Apostle Paul plainly testifies that the things written were written for our learning, so that through the patience and consolation of the scriptures we may have hope; therefore, we must also apply these things to our present circumstances. For Christ, the Lord of all, when he foresaw how greatly Satan would, through his chosen members, the old and new Roman Empire, afflict the church, would have us diligently admonished of every thing, so that all afflicted persons should learn patience from this, and conceive comfort and hope, and not be discouraged by the heavy burden of evils. Just as he has therefore diligently described the old Roman Empire, and showed as it were pointing with the finger what mischief it should work to the church, and admonished all to have faith and patience; so will he from hence forth describe papistry or Antichristianism, in which description he sets forth before our eyes whatever the saints shall suffer; that being warned beforehand, they may endure persecution more manfully and yield less to misfortunes.

\index{Antichrist}\index{Jerome, Saint}\index{Daniel (Book of)}\index{Rome}And in goodly order he begins to set forth Antichrist after the Roman Empire is torn and taken away. For Daniel says that a little horn should arise among the ten horns, and that three of those horns should be pulled down, plucked off, and cast away, so as to attain to great power. For he signifies that the Roman Empire, being divided and brought now to decay, Antichrist shall arise, who will procure to himself a new and counterfeit empire. And Paul says also that Christ shall not come to judgment until Antichrist has gone before; and that he shall not come neither unless this be first taken away, which hinders and prevents his coming. The which Jerome and other holy expositors understand of the Roman Empire, which must be plucked up and taken away, and then Antichrist shall arise. But the majesty of the empire was destroyed about the year of our Lord 480, when Odacer invaded Rome. For from that time, for the space of 300 years and more, there was no emperor of the West after Augustulus. And besides this, under the emperor Justinian, Rome was burned and laid waste by Totila. Since which time the bishops of Rome have begun to look aloft and to think upon a new kingdom.

And therefore the Lord says that this beast arises from the very earth. The kingdom of our Lord Jesus Christ comes from heaven and brings things to heaven. Papistry comes neither from Christ nor from his doctrine, but arises from the earth: that is to say, from evil means, ambition, avarice, treason, and cruelty. What ministers of the church Christ ordained is easily perceived by the Gospel of Jesus Christ. That he forbade them government, supremacy, authority, and majority (as they term it) appears in the 18th and 20th chapters of Matthew and the 22nd chapter of Luke. Therefore, the Acts of the Apostles and the doctrine of Peter testify that Peter was a minister, and not a lord of the Apostles, much less a prince of the city or empire of Rome. For they lie boldly that say how Rome and Italy are the patrimony of Saint Peter, given him by the Lord. At first, the Apostles and apostolic men, ministers of churches, governed the churches equally, neither did one take upon himself more preeminence than another. Which thing I am able to prove by many testimonies of ancient writers, if need required. About the time of the Council of Nice, and a little before that time, when churches were greatly multiplied, Metropolitans were ordained and customarily received, instituted indeed by a laudable [?but yet man] ordinance: that is to say, in a certain province or head city a bishop or pastor was ordained, who should have as it were the oversight of the rest, and should serve for the calling of Synods or assemblies. Yet it was then diligently provided that he should not be called Primate, lest any man should think himself preferred before others in power, but in order. Neither was the bishop of Rome at that time exalted above all others, but there were diverse Metropolitans, whereof the bishop of Rome was one. The Council of Nice confirmed that same custom and would have it ratified. Socrates in his Ecclesiastical History, book 5, chapter 8, recounts many Metropolitan churches in Asia. Saint Jerome to Evagrius and in an epistle to Titus says plainly that in old time churches were governed by the common counsel of priests or elders, and that then bishops and priests were all one: After, by the custom of the church, not of the verity of the Lord’s ordinance (I rehearse Saint Jerome’s words), bishops were preferred before priests, yet must they govern churches together.

\index{Augustine, Saint}And of that same custom, yea rather of the abuse of the custom, Antichrist had his beginning. For Boniface, Bishop of Rome, began first to assume dominion over the churches of Africa. But he was immediately restrained by the six African councils, where Saint Augustine is also said to have been present. After that, the Bishop of Constantinople also began to claim supremacy for this reason chiefly, that Constantinople was then the imperial palace and chief city of the empire. However, certain bishops resisted him; among whom was Leo, Bishop of old Rome. There remain certain letters of his to the Emperor of Constantinople, to the bishops of the East, and to others. So was this trouble appeased for that time also. But straightway another Bishop of Constantinople, blinded with ambition, demanded anew to have the supremacy given him. Whom Pelagius and Gregory, Bishops of Rome, withstood. And this latter so impugned the supremacy of the Patriarch of Constantinople that he did not hesitate to call him the vaunter of Antichrist, who would usurp the title of general bishop. There remain not a few letters written of this matter in his register.

\index{Papacy}Nevertheless, a few years later, when the bishops of Rome were greatly afraid that that dignity would be given to the bishops of Constantinople, Boniface III obtained from the emperor Phocas—a parricide—that the bishop of old Rome might be considered the universal bishop, and Rome the head of all churches. This constitution established the Pope in authority, so that he was now regarded by many western bishops as Apostolic, and many matters were brought before him for judgment. Through this, he gained the favor of many princes, chiefly of France, by whose aid he drove out of Italy both the emperor of Greece and the kings of Lombardy, and brought Rome and the best and most flourishing parts of Italy under his own dominion. Thus I say, out of the earth comes up the second beast.

Furthermore, Christ calls the Roman papistry a beast, because in avarice, covetousness, tyranny, cruelty, and even in bestial behavior, it differs nothing from the old beast, of whom I have spoken before.

\index{John, Saint}Up to this point, we have discussed the origin of the Antichrist or Pope, and of the new empire. Furthermore, Saint John proceeds to describe that second beast vividly, so that we all may know and avoid it. And first, he reasons about the power of the Antichrist.

That other beast, he says, had two horns: and he adds, like a lamb. For of them is spoken in the fifteenth chapter of this book. And the Lord signifies the priesthood and kingdom, which the Popes usurp for themselves, affirming that power is given them in Heaven and on Earth, in spiritual matters and temporal. For therefore they give in their arms two keys, that is to say two horns. They boast that they have two swords. Of these blasphemies, he who will be fully instructed, let him read the words of the beasts of Boniface VIII, in the sixth Decree of Majorities and Obedience. One, in Clement V, second book of oaths; finally, Gregory IX, or rather the first book of Innocent III, title on Majorities and Obedience. All histories make mention that Boniface VIII died in the year of our Lord 1300, instituted the first Jubilee, and in the same openly before the people showed in the way of ostentation the Pontifical and Imperial majesty, while on one day he appeared in the apparel of a Bishop, on the other having put on purple robes showed himself to the people like an Emperor. They carried before him two swords. And he himself cried, “Lo, here are two swords,” as though he should point with his finger to the whole world, that he and certain of his predecessors and all his successors were that two-horned beast. What shall we say that all bishops by him consecrated wear upon their heads miters or two-horned caps? Unless therefore we be blinder than was Tiresias, we see with our eyes who is that great Antichrist.

And here we must observe that he does not say that those are the horns of a lamb. For Christ still keeps both the priesthood and kingdom with the faithful in the church; neither does he resign them to any other, he has appointed no Vicar. For he continually executes, at the right hand of the Father, the offices of both King and Bishop, and this all faithful feel with joy. He says therefore, like a lamb. For the Pope will make all men believe that he has received from Christ priesthood and empire, that he is Christ’s Vicar; where he is nothing less. He boasts everywhere that he is the great shepherd and has received the keys of the Kingdom of Heaven; and that from the very Lamb of God, through the Apostle Saint Peter; and therefore that all bishops are subject to him, finally, all kings, princes, and people.

He proceeds to show moreover, what the talk of Antichrist is, what is his doctrine, and what is his speech. He spoke, says he, as did the Dragon. The Dragon is the Devil, as was shown plainly before. Therefore, he ascribes to Antichrist, or Popery, a diabolical doctrine, or a devilish mouth or tongue. We must see, therefore, how the devil speaks, that we may understand rightly how Antichrist speaks. In Paradise, he so tempers his talk that he calls into doubt the certainty and truth of God’s word, and by that occasion places his own word instead of the word of God. Is it so, says the Devil, under peril of your life, has God forbidden you to eat of the fruit of the tree of knowledge of good and evil? Yea rather, if you eat thereof, you shall be made like unto God. And after the same sort, Antichrist in his Popery brings the truth of the Scripture into doubt, which by all means possible he discredits as imperfect, maimed, obscure, and doubtful. And immediately upon that occasion, he brings in his traditions and decrees, with which he may patch up what he contends lacks in the Scriptures. But in his traditions, he affirms things contrary to God’s word and so deceives men. And all men know, who have any skill of Popish matters, that the first and chief principle and foundation of Papistry is that the Scriptures are imperfect and obscure, and therefore need traditions. Moreover, the Dragon speaks openly against the laws of God, and so does the Pope manifestly. God will be worshipped alone; the Pope adds saints to him. God forbids idols and idolatry; the Pope commands them plainly. God will have his name to be sanctified, and to be sworn by only; the Pope, by dispensing with oaths, pollutes the name of the Lord and commands us to swear by the names of God. God commands us to keep holy the Sabbath day; the Pope brings this into contempt, sets forth his own holy days, and makes double feasts. God commands us to honor our parents; this the Pope abrogates and commands greater regard for abbots and abbesses. God commands, “You shall not kill, you shall not commit adultery, or steal.” The Pope grants most ample indulgences and pardons to his soldiers for rash wars made at his will and pleasure; he spoils with his sacrileges all churches; and he, with his maiden priests, fills all the world with adulteries and whoredoms, to speak at this time of nothing else more filthy. And where God forbids lies and false witness, the Pope and his whole doctrine (which he sets forth besides the Scripture) are sown with lies. He not only dispenses with false witnesses but permits also to break safe conduits and public faith given, and if he hates the Prince, absolves the subjects from their fidelity and obedience; gives liberty to all lusts, and makes laws which nourish the desires of the flesh. And which shall be the Dragon’s mouth, if this is not it? The Dragon moreover is said to have spoken and said to the Lord: “All these things I will give thee (for he showed him the kingdoms of the world) if thou wilt fall down and worship me.” What other thing speaks the Pope? Does he not enrich his obedient children with the riches of this world, especially those who will fall down and kiss his feet? I suppose the Devil would never be so shameless as to offer to the Lord his foot to kiss; but that beast, in the sight of God and his angels, and of all the world, dares put out his foot, marked (not without great mockery) with the sign of the Cross, and proffer it to be kissed by all the children of God. I cannot bring forth the horrible and innumerable blasphemies out of the decrees and decretals. For I am ashamed of such ungodliness. Who therefore will not acknowledge that Satan himself does in this beast reign and rage? God shortly confound the same. Amen. Amen.
