\chapter[Sermon 86. The Description of the Judgment, Wherein Punishment Is Taken of Antichristians and Ungodly.]{The Description of the Judgment, Wherein Punishment Is Taken of Antichristians and Ungodly.}
\label{sermon:086}
\markright{Sermon 86. The Description of the Judgment, Wherein Punishment Is ...}

And I saw an angel standing in the sun, and he cried with a loud voice, saying to all the birds that fly in the middle of the sky: “Come and gather yourselves together to the supper of the great God, that you may eat the flesh of kings, and of high captains, and the flesh of mighty men, and the flesh of horses, and of those who sit on them, and the flesh of all free people and bond people, both small and great.” And I saw the beast and the kings of the earth, and their armies gathered together, to wage battle against him who sat on the horse, and against his soldiers. And the beast was seized, along with that false prophet who had worked miracles before him, with which he had deceived those who received the beast’s mark and those who worshiped his image. These two were cast alive into a pool of fire burning with brimstone, and the remnant were slain with the sword of him who sat on the horse, a sword that proceeded from his mouth, and all the birds were filled with their flesh.

\index{Judgment, Last}\index{Ezekiel (Book of)}Immediately after the description of the Judge, and a vivid depiction, a description no less evident follows of the judgment: that is to say, how Christ, having vanquished his enemies, commits them to perpetual torments. And the Apostle uses a prophetic phrase and eloquence. For by a figurative speech, all fowls are called to the slaughter and feast, that they might be filled with the flesh of the slain. And first an allusion is made of such as slay wares and prepare a feast, whereunto they may call their friends, and make them cheer. Again an allusion is made to the murder and slaughter of enemies, whereof wild beasts and ravening fowls are filled. Neither is there any thing hereby signified, but Christ shall overthrow all the ungodly, and take punishment of the same. Before was set forth a supper for the godly, wherewith they are refreshed and fulfilled. Now is prepared a feast of the solemn slaughter, whereby the ungodly receive no commodity, neither are they satisfied, but rather are slain and devoured, that is to say, perish. For no man will imagine that the wicked shall be overthrown at once, and after wearied by wild beasts, and gnawed by fowls, and so all punishment gathered together. For so should their pain seem to be none at all. But by temporal parables, eternal things are figured. These are taken out of the Prophets, namely out of the sixty-sixth chapter of Isaiah, and the thirty-ninth of Ezekiel, where are read in a manner the same words when he covers and wraps the ungodly with calamities, I mean wars, and destructions, and with other torments as it were kills them: but chiefly, when at the last judgment he commits them to pains everlasting.

And this slaughter is declared by the angel, standing in the sun and crying with a loud voice. This signifies that the day of judgment will be solemnly proclaimed with great fanfare and will be a remarkable day, so that no one can be ignorant of it, but all things concerning it will be heard by all people. Therefore, he cries with a loud and audible voice. And he gathers, as in Ezekiel, the birds and fowls to devour kings, and people of all ranks, ages, and genders: that is to say, that all these must be called together to suffer eternal punishment and destruction. Therefore, he lists diligently kings, chief captains, strong men, and people of all sorts, namely, persecutors of Christ, Antichristians, the ungodly, scoffers, and impenitent persons.

\index{Papacy}\index{Rome}Here is shown the cause of the damnation and destruction of the ungodly, while their enterprise, endeavor, and attempt is shown. They are now assembled to fight a battle against Christ and his elect, that is, against the church, by the beast, the Kings of the Earth, and their armies, captains, and soldiers, fighting both spiritually and physically. And here no lengthy commentary is needed. Read the histories of the church for the span of these five hundred years. And see what is done today by Popes, Bishops, and Princes, and by their counselors and ministers. Parliaments are called, in them bloody decrees and laws are made against the evangelists. And a grievous persecution is attempted against the Gospel and church. The spiritual fathers cluster together, they consult, and cry out that a council must be called. Wherefore, I pray you? to the intent that the new doctrine (as they call it) of the Gospel may be cut up, rooted out, and plucked up by the roots. And therefore they stand always ready to fight, are many times assembled against God and his anointed, to fight with the Lord Christ and with his chosen. Now days, if at any time peace is concluded, and the most bloody wars of Princes are taken up, or in the composition of peace, or immediately after, consultations are had, how to oppress the godly. But there is no other kind of sin more heinous than to impugn the truth of the Gospel. And therefore this is the chief cause of the condemnation of the wicked at the same day. It is accounted at this day among the chiefest virtues of Princes, if a Prince will give no place to the preaching of the gospel, but will shut it out, defend and maintain the church of Rome, with those doctrines, rites, and ceremonies falsely called old. Such are called right and good Catholics, most Christian, and defenders of the faith.

But now their condemnation is certain, and the manner of it is described. And the beast is captured. He speaks of these well-known offenders, as if they were seized by the manner, and taken in fact beyond their expectation. For in the midst of their plans they are intercepted, while they are yet in great hope, and truly believe they will accomplish many things against Christ and his church; then, in the heat and wickedness of their attacks and persecution, they will be apprehended. Consequently, it is also clear that persecution of the truth will continue until the end of the world. But who will be taken? The beast and the false prophet, who performed miracles. Concerning these matters, see what is said at the end of chapter 13. The cause of eternal damnation is again presented here. For he has deceived the world by his enchantments and crafty trickery, by his decrees and commandments. Of this I have spoken before, more than once, as also in chapter 17 and following verses. The fellowship of his condemnation is also joined: all those who have received the mark of the beast and have worshipped the beast. Concerning this, we will now repeat nothing. These things are declared in chapter 13. And although nothing can or ought to be proven by the pictures, it is nevertheless certain that painters have borrowed from this source for their old paintings of the last judgment.

We see the old depictions of the final judgment, painted a hundred years and more, intended to represent and display to us a large crowd of priests, monks, and friars, and all kinds of spiritual leaders. Most notably, the spirits of kings and popes are shown being driven to hell, where they burn in its bottom with everlasting fire. Consequently, it is said that more priests than farmers go to the devil, and so forth.

Furthermore, the damnation itself, and the manner of torment, or the torment itself, shall be fire. For John says: these two are cast quickly into a pool burning with brimstone into a lake or standing water as is found in fen countries, for such is the description of hell and of extreme punishment, as is also described by Isaiah in chapter 30. Tophet was long ago prepared, and the same is prepared for the king, which he has made both wide and deep; the inner chamber of it is fire, a great store of wood, which the blast of the Lord, or stream of sulfur, sets on fire. Not dissimilar things are read in chapter 66 and in the gospel of Jesus Christ, Mark 9, Matthew 25, and in other places without number. And the plague of Sodom is known to all men, Genesis 19. The laughings therefore of the ungodly scoffers, the godly would rather believe these things than prove them. For here is hell set open as it were for us to look into it. Let us fear.

\index{Antichrist}And two here are singularly named. For sins, they have been authors of all evils, and rightly must be chief in pains or torments. For the wise man also has said, terribly and surely he will appear to you, for it shall be a most severe judgment to those who bear rule, and the mighty shall suffer mighty torments. For Scripture also shows in another place that there are degrees of punishment, according to the quality of the crime. And let us not think here that the head is punished without the members. For the whole body of Antichrist shall be condemned to torments. All the ungodly shall be punished, as hereafter shall be more plainly declared, and in the end of the twentieth chapter.

\index{Resurrection}But this also is especially to be noted, that it is said how they shall be cast quick into hell. For so is signified the resurrection of the dead. Here is signified that in the judgment the world yet remaining shall be taken in the flesh, not as yet dead but living, which Saint Paul expresses vividly. 1 Thessalonians 4. And we pronounce openly in the creed, saying: from thence he shall come to judge the quick and the dead: not only the just and unjust, but the dead, to wit remaining in the flesh, and living. Antichrist therefore shall live at the day of judgment, and shall not be extinguished before. The persecution of Antichrist shall endure, with all ungodliness, even to the last day. And just as Korah, Dathan, and Abiron, and the rest of the conspirators, were taken in the very crime of rebellion, and swallowed up quick with their tabernacles, and all their things of the earth opening, so at the day of judgment hell gaping wide shall receive and swallow up all the ungodly, but chiefly the Antichristians. The which many now do not believe, but in that day shall find it with unspeakable pains, and horror incredible, and all we shall see it with these our eyes. Every one puts trust in his own sect, and hopes to obtain salvation in his superstition. But the things that we here at this present are told us of the judge himself, Christ, as most certain and undoubted, and after a sort are set forth to behold.

And what shall be done at that judgment with the remainder of the ungodly and impenitent? Shall only Antichristians be damned for Antichristianism? John adds: and the remnant are slain with the sword, and so forth. For in Matthew 25, the judge declares: “Go into everlasting fire, prepared for the devil and his angels.” For I was hungry, and they gave me no food, and so forth. For if they are to be damned by the just judgment of God, who, when they might have done good to men, have not done it, what, I pray thee, shall become of those who not only have shown no generosity to the needy, but have moreover plundered those who had honestly and were generous of their goods, and afterward spent them in riotous living? And so they have brought those of honest substance into extreme misery, and by this means have robbed the poor also, who were wont to be helped through their generosity, of their help and succour? Here are also comprehended heretics, Jews, Gentiles, Mahometans, and all others like them.

\index{Primasius}At the end, it is repeated that all birds are filled with the flesh of the damned. We understand this to be expressed through figurative language, and it should not be interpreted literally, but rather as signifying that all ungodly and impenitent people will be punished most abundantly. Primasius, interpreting this passage, says that we should not understand it carnally, believing that the saints—which he explains as being represented by birds—are satiated with the flesh of the wicked. Instead, the equity of God’s judgment, revealed to the saints, shows how, in redeeming the full number of the elect, the remainder is decreed to be damned. They are said to be filled with this knowledge of righteousness, which a person may hunger or thirst for in this life but cannot fully comprehend. Isaiah also speaks of the ungodly: “And they shall be for the fullness of sight unto all flesh.” Here, I suppose, is the fullness of souls mentioned earlier. Immediately following, “and the fowls may be taken in the evil part,” referring to the angels who transgressed. After leading their followers to destruction and accomplishing their evil desires, they are said to be filled with the flesh of the condemned, taking satisfaction from their damnation, to which they were authors of errors. This is what Primasius says. However, while I do not disagree with these things, which are undoubtedly spoken truthfully, I believe they should not be examined so closely. What is spoken figuratively and in the prophetic manner seems to indicate nothing other than that all the ungodly will be destroyed by the great power of God and extreme torments. Therefore, let us fear God, to whom alone be glory.
