\chapter[Sermon 48. Of the Cruel Fight of Antichrist Against the Prophets of God, Whom He Overcomes and Slays, and Shamefully Uses Them.]{Of the Cruel Fight of Antichrist Against the Prophets of God, Whom He Overcomes and Slays, and Shamefully Uses Them.}
\label{sermon:048}
\markright{Sermon 48. Of the Cruel Fight of Antichrist Against the Prophets ...}

And when they have finished their testimony, the beast that came out of the bottomless pit shall make war against them and overcome them, killing them. And their bodies shall lie in the streets of the great city, which spiritually is called Sodom and Egypt, where our Lord was crucified. And some of the people, kindreds, tongues, and nations shall see their bodies for three and a half days and shall not allow their bodies to be put in graves. And those who dwell upon the Earth shall rejoice over them and be glad, and shall send gifts one to another, for these two prophets vexed those who dwell on the Earth.

\index{Antichrist}\index{John, Saint}\index{Rome}We have heard of the constant preaching of those who will oppose Antichrist and his army for Christ’s truth and the church of the faithful, and that throughout the time Antichrist exercises tyranny against the church. Consequently, our Lord Jesus Christ will teach us through the Apostle and Evangelist Saint John, concerning the condition in which the saints will fight and how Antichrist will encounter them; which also pertains to consolation and a necessary admonition, lest any man should be discouraged by the prosperity of the Antichristians and the calamities of the faithful. He speaks therefore expressly of the grievous persecution of Antichrist, which has now continued these many years—I mean all that time wherein the Bishop of Rome has usurped authority over all churches—with some small spaces of respite permitted by the Lord. This persecution of Antichrist is more grievous and longer than any that occurred among the ancient people of God or in the primitive church. Certainly, for these fifty years, whoever, of whatever state or condition, began to speak even a little against the church of Rome, immediately felt hatred, imprisonment, banishment, and death. History testifies to this, and shows also that the persecution increased as the bishops themselves and their champions, monks and friars, increased in number and power.

And the Lord declares most diligently when, who, of what status, where, when, and with how great cruelty the Antichrist shall act as a tyrant against the faithful servants of God. He adds immediately that all his endeavors shall be utterly vain; and how great shall be the rewards of steadfast ministers, and also the calamities of the Antichristians.

\index{Arethas of Caesarea}First, indeed, he plainly admonishes when persecution must arise: not before the testimony of the prophets is finished. I showed you before that the testimony is the sincere preaching of the gospel. Arethas asks: what testimony? That the one who will be present is not Christ, but a deceiver and a pestilent seducer, and so on. And so great is God’s goodness, loving his church, that he will not allow the preachers to be taken away until they have finished their preaching. For the gospel must be openly preached to all men for salvation, and deliverance from anguish, schemes, and deceits, and from the seducers of Antichrist. And they shall finish their ministry with numerous writings and continual preaching. They shall finish, I say, when it pleases God. For some preach and abide safe and sound for many years, being safe and secure from persecutions. And others are immediately apprehended, cast in prison, and slain. Thus are these things done, as it seems good to God; which must ever be credited, whatever means he uses, to advance his glory and further the health of his church. Here comes to pass also, as we read often in the gospel, that the Lord was not taken, because his hour was not come. Therefore, a certain hour also will be appointed by God to the preachers. Before this hour they are safe and secure, though the devil be never so furious, tyrants rage, and bloodsuckers and enemies of faith lie in wait. We marvel sometimes how the preachers of the gospel could preach in so great a company of wolves for so long a time, and directly against wolves. Why were they not immediately torn to pieces?

\index{Luther, Martin}The Lord God almighty has kept them, so that they might thoroughly finish the testimony of the truth. He allowed their enemies and gave strength to his servants to preach. To him we shall render thanks that many good preachers in times past, and of late days Doctor Luther, and Doctor Zwinglius, and other faithful witnesses of God, could in so wicked a world and in so great a power of Antichrist, execute their ministry for so many years, in defiance of the gates of Hell. Nevertheless, the princes and magistrates deserve also to be praised for the lawful defense shown them; yet this should have been none at all, unless the power of God would have had it so.

\index{Papacy}And when the faithful in the church have been sufficiently warned, so that those who desire to be wise and are not of a steadfast purpose may all escape the snares of Antichrist and live in Christ, persecution will immediately follow. For as soon as the Pope hears, with his supporters, that he is being attacked, he will immediately begin to threaten and unleash his power, ultimately seeking to incite secular authority against heretics. He explicitly shows who this enemy of these prophets and their preaching will be, namely the beast, which is the Bishop of Rome, recognizable by his cruel, tyrannical, and beastly power. More will be said about the beast in chapters 13 and 17, where we will hear that she arises from the earth, from the bottomless pit, and from the deep pit of hell. For the origin of that wickedness is attributed to no other source than the devil, the prince of hell, a liar and murderer. And the thing itself proclaims this today: all persecutions and conflicts are instigated by the Pope and his bloody ministers of evil. From such authors arose all the calamities of former times.

\index{Jerome, Saint}And he contends against the ministers and ministry of Christ with sophistry, with crafty and subtle practices, excommunication, death, and terrors. Jerome says: Antichrist will inflict various kinds of torments, and those he cannot overcome, he will attempt to vanquish with doctrine. He will give rewards and will promise sweet words, and will also display false miracles, and so forth. And saying it has pleased the Lord to call that seat the beast, why should we call it the holy See? If the Pope is that bloody beast, why should we address him as most holy father?

\index{Daniel (Book of)}He shows moreover, with what fortune Antichrist shall contend with the prophets. He shall overcome, says he, and kill them. The same the Lord said plainly in the Gospel, Matthew 10, and John 16. And before also Daniel in chapters 7 and 11. Some things are spoken also before in this book of the holy martyrs. The Lord gives this warning in time, lest if we should see the preachers of the evangelical truth slain, we should doubt of the truth of the preaching, or esteem the matter of religion after the felicity of this world. Which nevertheless many do at this day. For most men say: if this were the preaching of the truth, as it is said to be, the most true God would not forsake his own cause. But now since the preachers are oppressed and destroyed, why should we not gather that their matter is false, and theirs true that overcome? But if we might so reason, then the Prophets, Christ, and the Apostles defended a very evil cause. For all in a manner being oppressed of their enemies, in the end were slain also. Full good then was the quarrel of the Jews, Pharisees, and the most wicked enemies of God’s word? How be it, you will say, since the truth is invincible, how is Antichrist said to overcome? He shall not overcome doubtless, by sure testimonies, by holy Scriptures, or strong reasons: but by force, imprisonment, sword, and fire. For therefore by interpretation immediately follows: and he will kill them. Therefore by killing he shall seem a conqueror. For if in a combat Aeneas shall overcome and slay Turnus, Aeneas shall be called a victor. And hitherto in deed Antichrist overcomes: and although the martyrs be slain, yet do they before God receive the reward of victors: because their cause is just, and the truth overcomes in them. The enemies overcome with the multitude, pomp, authority, power, favor, riches and other like things: we in the goodness, and excellency of the cause, and finally by better testimonies of the Prophets and Apostles.

Now we have the manner of the conflict and victory. He will fight and overcome with physical weapons, and will be subdued with spiritual armor. Added to this is what cruelty he will use against the prophets. He expresses this in two statements: “Their bodies will lie in the streets of the city, and they will not allow their bodies to be placed in graves.”

The first sentence indicates extreme cruelty combined with utter contempt. For all foul things are cast out into the street, even the refuse of all streets is trampled underfoot. Antichrist will therefore treat the prophets with great shame, so that all will believe they have power over them, and will, as it were, spurn them with their feet, considering them outcasts and wicked persons, whose removal will make all things safe. Indeed, in some cities, it is customary to cast the bodies of those who are executed into the streets, so that all may tread upon them and drive carts over them, for the terror of others, and to signify that those executed were most detestable and put to death for serious crimes. And to this pertains that Antichrist, by secular power, will publicly hang up some ministers of the church on the gallows, and fasten others with chains to a post, and then burn them with a slow fire, and finally kill them, binding them so terribly to the pale in chains that he may hoist them up and let them down into the fire again, singeing them and lifting them up again, for the terror of all who look on. What will they say about him, judging them unworthy of the last honor? Burial is the last honor done to a person, but he will not allow the bodies of the faithful to be buried. Thus, perhaps, he explains what he said before, that their bodies will lie in the streets. At this day, not only burial is denied to those who suffer for the gospel, but also the bones of the dead are dug up and burned, [?those] who, while alive, would not receive the papal sacraments. For if anyone dies without having whispered a confession to a priest, confessing all his sins to him, without asking for absolution or receiving the Lord’s bread, or allowing himself to be anointed with extreme unction, although he died in true faith, because he has not used those papal ceremonies and subjected himself to the Pope, he will not receive Christian burial, but will be buried on a dung heap with dogs. The practice itself demonstrates this today. Moreover, these Antichrists will seek by this means to abolish all memory of the godly. For monuments are made to preserve the honorable memory of the dead. But the righteous will be in eternal memory, and so on. And they indeed think that they are acting like good Catholics, but the Lord Jesus exposes their work, saying that it is extreme cruelty. Then what shall you think of those who, blinded by hatred of the true religion, like wolves and ravens, swoop down on the bodies of dead martyrs, tear them apart, and treat them with great shame?

But that cruelty is harsh and bloody, and it shall continue for three and a half days, a period that all commentators understand to be brief, certainly so, but yet uncertain, as I mentioned previously regarding months and years. Therefore, I suppose this shortness of time is brought about for consolation. We also say that the Lord grants the afflicted spaces to breathe, and shortens the sorrowful days, so that we might be able to endure it. If, therefore, our patience is tested in a severe persecution by Antichrist, let us think that our Lord God has accounted for all the days of our calamity, and that he has shortened them for the consolation of the weak.

And the very place where this cruelty must be wrought against the prophets, he expresses plainly, and as it were points out with the finger. To wit, the great city. And it is the city of God, and is also the city of the devil: it is the city of Abel, the innocent, it is also the city of Cain, the parricide: it is the catholic city of saints, it is also the synagogue of Antichrist. These cities are open throughout the whole world, and are enclosed with no straight walls. You might call this city the lordship, dominion, kingdom, or empire, or fellowship of the wicked. Wherever therefore Antichrist or the Pope of Rome has jurisdiction, and even in the Roman church itself, throughout all nations and people, these things which we have heard shall be done against martyrs. For setting forth that city with more plain tokens, their bodies, he says, shall lie in the streets of the great city. And by explanation he adds: which is called spiritually Sodom and Egypt. Moreover, where also our Lord was crucified. And again: and they shall see of people, and kindred, and tongues and nations; therefore by this he understood not any straight, nor yet any large city enclosed with walls, but that city stretches throughout the world wherein dwell nations, kindred, people, and so on. Sodom and Egypt are far apart, neither can they be joined together with any walls. Again, our Lord was crucified in the city of Jerusalem, which is also called by the Prophets Sodom and Gomorrah, but he is crucified daily in his members throughout the world. And there is one and the same city and society of all the wicked in the world, as there is one body of the godly. Let us know therefore how that city wherein the bodies of the prophets lie in the streets, is the city of Cain, and the Roman church scattered over the world. The same is called Egypt and Sodom, but spiritually.

\index{Ezekiel (Book of)}Where we see this vocable [illegible] spiritually, to be used in a sense far from the letter, for otherwise there is no spirit at all, either in Sodom, or in Egypt: for they are altogether flesh. Therefore, in sense of a parable, and by comparison, this city is called Sodom and Egypt. What Sodom was, appears in the 19th chapter of Genesis and the 16th of Ezekiel. Her sins went up to heaven. But what manner one is at this day, and has been for a long time, the church of Rome, all men know, except it be they that will not know. And the Apostle in 1 to the Romans has expounded. And Egypt robbed the children of God of their liberty, oppressed them with vile bondage, and prohibited them from the true worship of God. So likewise the Roman church has spoiled the church of Christ of the liberty gotten by Christ, has wrapped her in filthy servitude, that she might serve in the dirt of men's traditions. It prohibits moreover by all force and power, that she shall not, in returning to the gospel, serve God truly. Verily, our Lord Jesus Christ was crucified on Mount Calvary, at the city of Jerusalem; and also the articles of our faith expounding the same, say, that he suffered under Pontius Pilate. He was the Roman governor, it is manifest therefore, that Christ suffered under the Roman Empire. Under the same Empire, and under the judgment thereof, were executed the Apostles and ancient martyrs. Under the Empire of new Rome fall those prophets also at this day by sword and fire. Also, people, kindred, tongues and nations obey this Empire, now called the church, so that preachers in all places, the beast so willing and commanding, are apprehended, and slain with cruel deaths, people, kindred and nations looking on.

To amplify and most aptly express the cruelty, this pertains chiefly, that these earthly men (possessing in these lands a church entirely carnal) shall rejoice and be glad over the calamities and miserable deaths of prophets. The Lord himself prophesied this before in John 16. Verily, verily, I say unto you, you shall weep and lament, but the world shall rejoice, and so forth. Yea, they shall send gifts, he says, and letters of rejoicing. That this was done in the council of Constance, when John Hus and Jerome of Prague were burned, histories make mention. We have heard very lately, how after England fell again to the Roman religion, what joy and gladness, what banquets and triumphs the Papists made in all places. So often as the ministers, or other faithful are burned, they keep a festive air and pleasant feasting, singing “Te deum Laudamus.” And letters of rejoicing fly to and fro; in some other places, with solemn procession, they rejoice at the misery of the faithful, but the Lord sees these things, which foretold sins that the same things should now come to pass.

And the cause of this great joy is no other than that those prophets have troubled those who live on earth. For those who love the earth and desire earthly things are greatly offended by the free preaching of the truth, which they hate more than dogs and snakes. For they either seek to attain honor, riches, and pleasure, or, if they have them, they desire to keep them. But they are greatly afraid that through preaching, these things will be severely shaken or entirely taken from them. Therefore, they desire nothing more than to be rid of and delivered from their clamors, and to have them removed immediately. For so they think they will be safe and enjoy their pleasures at will. With like affection and counsel as in Herod’s feast, they coveted not a kingdom nor a great sum of gold, but the head of John the Baptist. The Popes would rather at this day have the heads of certain ministers of the church than so many thousand crowns. Yes, moreover, ministers of churches are called plagues, disturbers, seditious, and injurious against God and his saints, and against all men. Therefore, they wish with all their hearts to be eased of this burden. The Lord Jesus forgive them this sin.
