\chapter[Sermon 33. The Sixth Seal Is Opened, and the Corrupting of the Sincere Doctrine Is Exhibited.]{The Sixth Seal Is Opened, and the Corrupting of the Sincere Doctrine Is Exhibited.}
\label{sermon:033}
\markright{Sermon 33. The Sixth Seal Is Opened, and the Corrupting of the ...}

And I saw, when he had opened the sixth seal, and behold there was a great earthquake. And the sun was as black as a sackcloth made of hair: and the moon became as blood; and the stars of Heaven fell to the earth, even as a fig tree casts its figs when it is shaken by a mighty wind: and Heaven vanished away as a scroll when it is rolled together.

The opening of the sixth seal of the lamb generally reveals to us a scene visible to all men, depicting the corruption of doctrine within the church, with a mournful and terrible effect. Nor is there anything said here in the sixth seal, as also in the five preceding seals, other than what our Lord Jesus Christ foretold previously in Matthew 24: that the Gospel should be preached throughout the world, and that wars, famines, pestilences, and grievous persecutions would arise, along with false prophets who would deceive people and drown them in great sorrows.

Nevertheless, these things must be explained with religious care. For it is not to be thought, simply because the Lamb opens the sixth seal and the Son immediately appears dark, that Christ is the author of corrupt and evil doctrine. For Christ is the one who sows good seed in the field; the enemy sows darnel, as the Lord himself explains in Matthew 13. Christ teaches sound doctrine through the Apostles and sincere preachers; and when it seems vile to the world and cannot please, by his just judgment he leaves the contemners to their affections, as the Apostle Saint Paul says, he sends upon them the power of delusion, that they may believe lies, and so be judged—all those who would rather believe a lie than the truth. And the deception through corrupt doctrine is a more harmful evil than the bloody persecutions. Indeed, the seducers and false prophets have done more harm to the church than cruel tyrants. Finally, people are more grievously punished when they are abandoned to be seduced by deceivers than when they are subjected to being torn into pieces by their murderers. Therefore, it is a most grievous plague of God, and utterly to be abhorred, when the simple truth is despised and delivered to lying deceivers, who, it may be spoken with reverence, seem to all to be [illegible] and be peace. For where the gospel is purely preached to many, these men say, “I do not understand what these men teach us from the Gospel; but I can see that the old [?have] all to be raided, and these new both to be peace and be[illegible].” Therefore, you shall have teachers who will perform to you in deed the same that you talk. Would God we lacked examples, and did not see certain nations that formerly had the free and pure preaching of the gospel, and now stripped of all truth, sigh under the pleasure and boldness of most wicked deceivers, who trample God’s word under foot, condemn it as heresy, and stop the mouths of the wretched people full of human dung. This is the punishment of the truth despised.

\index{Antichrist}\index{Mahomet}\index{Judgment, Last}And this passage may not be explained as referring to a single age, for the sins mentioned are things rehearsed in general; but rather to the entire time stretching from the age of the Apostles to the last judgment. It contains therefore the corrupt doctrine of Valentine, Marcion, Manichaeus, Arius, Macedonius, Nestorius, Eutyches, Donatus, Pelagius, the followers of Priscilla, and finally of all heretics, and the confused mixture of Mahomet composed of the same, and chiefly the sophistry and most corrupt doctrine of Antichrist and of his ministers.

But when the Lamb opened the sixth seal, there was not heard now as before the voice of the Beasts, Elders, or Martyrs, but a terrible earthquake. An earthquake in the Scriptures signifies a wonderful commotion of all things, troubles, tumults, and great alterations. And truly, greater darkness arises from nothing else than by altering godly religion and receiving of wicked doctrine. For so arise sects, seditions, and wars. You may see many examples of this in the history of the ancient people, who are recorded to have been grievously shaken so often as they have changed their religion and kind of doctrine. By this earthquake, therefore, is signified that exceedingly great trouble shall arise from this, for that a new and a strange kind of doctrine should be brought into the world, amiss inciting men.

Here is what you may say to those who blame the Gospel and its preachers for all the current troubles, seditions, and upheavals in the world. Elijah once made an answer for us that will suffice for all times; the passage is in the third book of Kings, chapter 18. “I have not troubled Israel, but you and your fathers’ house, who have forsaken God, \&c.” The story of Jeremiah in chapter 44 is also relevant, where all the evils then afflicting the wicked were wrongly attributed to sincere doctrine and to the prophet Jeremiah. Learn also what to say to those who claim that God has allowed his Church to lie and rot in errors for so many years [?].

And the corrupt doctrine is described by parts, even from the top to the toe, and the effect also of the corrupt doctrine is annexed. And first of all the sun, a planet most bright, not only waxes dark, but black also. And immediately is added an Image or a parable, like a hairy sack which is woven or made of hairs or of bristles. The sun lights and gives life to the world. And through Christ, which is the life of the world, we are illumined and quickened. He casts abroad from him the bright beams of the Evangelical truth. And just as Christ is not darkened in himself, so neither is the truth of the Gospel, which by nature is without pollution. By reason of the black clouds that override it, the light of the Sun waxes black and is hindered; and from the traditions of men, and the corrupting of the scripture arises darkness and blackness in matters of religion. The Gospel of itself is bright and wholesome. Christ is light, full redemption, health, and life most perfect. But when men had rather seek of others doctrine, life, and salvation, than of Christ and his wholesome Gospel, most thick and gross darkness arises in the minds of those men. For there is established another doctrine, righteousness, intercession, redemption, salvation, and life than that of Christ. They that receive that doctrine, seem to have put on them a shirt of hair which pricks, burns, and vexes continually. For there is no rest, quietness, security, or spiritual pleasure and refreshment from corrupt doctrine, but only tediousness. Christ pure, and sincerely received, is to man a joy unspeakable, and a most bright and joyful light.

After this, it is added that the whole moon, not a part only, is become bloody. For an image is again annexed, as blood. The moon receives light from the sun and is subject to courses, or changes, while one while it increases, and another while it decreases, and signifies the church. The church, established upon the rock, is not unstable; but because of its variable fortunes, is subject to many diverse chances. For now the church triumphs, and immediately being oppressed she mourns; now she increases in number, and immediately she is diminished. And the church is illuminated by Christ. But while the Sun itself is darkened, the moon cannot but be most obscure. Blood in the scriptures betokens great wickedness, chiefly idolatry and false worship of God. The Lord in Leviticus 17 says that he will account strange worship as blood. Therefore, when faith and knowledge are darkened in Christ’s church, it cannot be chosen but that blood shall arise in the universal church; that is to wit, the corrupt worship of God, which the Lord esteems as murder; there must needs innumerable sins and wickedness spring thereof. For the lively doctrine of Christ being corrupted, all things must of necessity be most corrupt, and swarm full of superstitions and iniquities.

\index{Daniel (Book of)}To these things is added another thing, which supports what has been said: stars fall from Heaven to the earth. Daniel called stars preachers in chapter 12. As also Saint Peter, in 2 Peter 2. Therefore, the preachers of churches depart from the heavenly doctrine of Christ, which was brought and revealed from heaven, and leads people to heaven, and keeps them in heavenly conversation; and they embrace earthly things, that is, the doctrine of men. By this, it comes to pass that both the sun is obscured and the moon is made bloody. Stars shine; preachers should proclaim Christ, the true light, to the whole world, but they have neglected this, being devoted to their own traditions.

To these is also added an image, which falls as the fig tree casts off its figs when shaken by a strong wind. Here is signified the corruption of preachers, and a great number of them. For the fig tree was made to produce sweet fruit; so also the ministry of the word was ordained for the salvation of men. However, the figs were unripe, therefore they remain green or untimely fruit. This signifies that the preachers were not ripe in true knowledge of Christ, and therefore are to be shaken down by every wind of doctrine, both that which they have admitted and that which they have proclaimed as earthly. The abundance of false teachers is signified by the fact that the unripe figs fall down in great abundance.

From these things now follows another, and Heaven departed, seeming to flee from human sight and vanish. Again, an image or likeness, like a scroll folded or rolled up, is added. Heaven in the Gospel signifies many times the kingdom of God. Therefore, the kingdom winds itself up on earth, and the church seems to hide herself, not that there will be no church at the last (for the church will always endure until the end of the world), but because, at the end of the world, the church will lie hidden, and it will not be considered the true church, which is indeed the true church. The letters and words are not wiped out of the book, but are not seen, rather they are hidden when it is rolled up. It is manifest today what John meant by this parable. For all, in a manner, judge the newly arisen Roman church to be the true church, which in truth is not the church of Christ; and the church which is the spouse of Christ is judged heretical. Therefore, the church is wrapped up and as rolled together. The Lord unfold and preserve it. Amen.
