\chapter[Sermon 41. The Sixth Trumpet Is Expounded, Wherein Is Treated of Saracens and Turkish Matters.]{The Sixth Trumpet Is Expounded, Wherein Is Treated of Saracens and Turkish Matters.}
\label{sermon:041}
\markright{Sermon 41. The Sixth Trumpet Is Expounded, Wherein Is Treated of ...}

One woe is past, and behold two woes come yet after this. And the sixth angel blew, and I heard a voice from the four corners of the golden altar, which is before the eyes of God, saying to the sixth angel who had the trumpet: “Loose the four angels who are bound in the great river Euphrates.” And the four angels were loosed, those who were prepared for an hour, a day, a month, and a year, to kill a third of mankind. And the number of the horsemen were twenty thousand times ten thousand. And I heard the number of them. And thus I saw the horses in a vision, and those who sat on them having fiery armor of a yellowish and brimstone-like color, and the heads of the horses were like the heads of lions. And out of their mouths went forth fire, smoke, and brimstone. And of these three, a third of mankind was killed—that is to say, by fire, smoke, and brimstone, which proceeded out of the mouths of them. For their power was in their mouths, and in their tails, for their tails were like unto serpents, having heads, and with them they hurt.

\index{Turks}The sixth conflict is that of Mahometanism, by the Saracens, Turks, and Tartars, fought with great cruelty and much suffering. And would God it were fought. For we perceive daily through the events themselves the mystery of the prophecy, and see its fulfillment, and even experience it as well.

At the sound of the trumpet of the sixth angel, John hears a voice from the four corners of the golden altar, that is to say from the midst of the altar: and there is no reason why we should seek a mystery in the fourth number. And he speaks of that altar which is before the eyes of God. That voice commands the angel trumpeter to release the four angels bound in the great river Euphrates. As soon as this was done, an innumerable army of horsemen marched forward, and slays and destroys a third of the earth, that is, a third of mankind. And the description of those horsemen, and their force or power, is most diligently given.

We have recently learned that the golden altar signifies Christ, sitting at the right hand of the Father. He is purer and more precious than gold; he is priest and sacrifice for all the faithful, standing before God’s eyes, that is, pleasing God, in whom or through whom his soul is pleased with all faithful, whose virtue is sufficient for all. And God the Father desires that such a one be preached and believed by all the faithful in the world. The ancient church, through the Apostles, believed and taught this same Christ; until, by the work and instigation of the Devil, most corrupt men arose in the church, some denying Christ’s deity, others his humanity, others separating the person consisting of God and man, and others confusing the natures or properties of the natures. God’s goodness patiently endured this assault, many times sending faithful and open defenders of the truth, who might root out these blasphemous errors—which we have read to be done by various bishops or preachers of the church, or by ecclesiastical assemblies, which we call councils: such as were the councils of Nicaea, Constantinople, Ephesus, and Chalcedon, in which Arius, Macedonius, Nestorius, Eutychius, and other monsters of heretics who opposed Christ were condemned. Nevertheless, the incurable perverseness of men gaining the upper hand, there was no end to alterations and blasphemies. For two great bishops of no small churches, Peter, patriarch of Antioch, and Severus of Constantinople, arose during the reign of Emperor Justinian, and impudently and most wickedly affirmed (as the acts of the Fifth Council of Constantinople amply declare) that the body of Christ was utterly incorruptible and truly deified, not subject to any affections as mortals are. For they asserted that the Word became flesh so that immediately it began to be one nature, namely, divine, that Christ was made [illegible], that is, incorruptible. These things seem to proceed from the most wicked school of Valentinus, Marcion, and Manichaeus. James of Syria, surnamed Zalos, of whom the Jacobites are named in the eastern lands, took it upon himself to defend the doctrine of Severus. He taught that Christ, because he was incorruptible, neither suffered nor was crucified, but that some other person suffered for Christ, while Christ only stood by invisibly and looked on. This opinion was refuted by many testimonies of Scripture and finally repulsed and overthrown with the articles of our faith.

\index{Antichrist}\index{Mahomet}For we profess in our belief that he who suffered under Pontius Pilate was crucified, dead, and buried. The prophets declared expressly beforehand that he should suffer and die; that he has suffered and died, the Apostles have witnessed, among whom John saw the death and suffering of Christ on the cross. Neither do we read that the Lord was ever so much offended with his disciples as he was with Peter, who sought to dissuade the suffering as unworthy of the Son of God. For he said, “Go behind Satan; you do not favor what is of God, but what is of me.” Therefore, they should not renew the error and madness that has been refuted. He seems to reason plausibly that God might have redeemed the world by another means than through the incarnation or suffering of the Son of God; to think it unworthy that the Son of God should be beaten with the blows of the wicked, and moreover slain. But this plausibility is from unclean flesh, not from God, indeed it is from Satan himself. Yet this absurd and most wicked opinion has found not a few followers. For the heresy of the Jacobites, contained in the Quran, is spread far and wide throughout the East. It is plain that the golden altar was by them most filthily polluted; and the merit of Christ’s suffering denied, the dignity and majesty of the priesthood and sacrifice of Christ trod under foot. There were, besides these, other most corrupt opinions in the West, and so forth. This thing worthily kindled the just wrath of God. For, in his just judgment, he permitted Muhammad to make new laws and to spread Jacobism far and wide throughout the world. For those who will not hear Christ worthily hear Antichrist, which thing the Apostle affirmed also in the first chapter to the Romans and the second chapter to the Thessalonians. Therefore, a voice is heard from the altar of him who sat on the right hand, commanding the four angels bound in the river Euphrates to be loosed, that is to say, to bring forth into the world destroyers who may overrun a great part of the world.

The heresies of the Nestorians, Jacobites, Monothelites are gaining force, as are monks and friars, and the Benedictine order of Cas-sina. Moreover, idols and images are increasingly present in the church, gaining strength, and the pride and unfaithfulness of the bishops are growing. Mahomet, the destroyer of the world, was born in Mecca, a city of Arabia, of very obscure parents. He was raised by Sergius, a vile monk polluted with all kinds of heresies. After he reached the age of twenty-five years, he claimed to be the prophet of God. Through sedition, he was driven out of Mecca, where a great rabble of Jews, Jacobites, Christians, pagans, and heretics had gathered. He then went to neighboring cities and houses, and for ten years secretly instilled his doctrine into the people, so that throughout Arabia there was found a great multitude of Mahomet’s sect. Then Homar, a bold man, taking about seventy other men ready to fight, asked Mahomet what he would do. He answered: “Truly, my will is that, executing the commandments of the law, you cleave to it in riches and poverty, and cleave together with mutual and steadfast love; that you do not defile other men’s wives through adultery; that you abstain from evil and prohibit others; that you do good yourselves and persuade others; that you make war in the name of God, and that by fear and force you set forth the laws to the disobedient; for these things truly I promise you Paradise.” At this talk, they gave each other their faith. Homar, with his sword drawn, swore that he would not suffer the preaching and law of Mahomet to be kept any longer secret. Thus, by preaching and by the sword, Mahometism prevailed greatly in a short space. They broke into Mecca, put down other religions, and beheaded those who resisted. There, this new Solon, Mahomet, proclaimed a new law in the Temple at Mecca. A great multitude of servants and wastrels resorted to that wicked fellow, who sent ambassadors to the people around him and solicited them to receive his religion, persuading many to wickedness. These things were done under Emperor Heraclius about the year of our Lord 620. And that wicked and most absurd law of Mahomet yet remains, and is called the Alcoran, so that it needs no further declaration. Nicephorus, in his history, says that the Saracens began the desolation of the whole world. Saracens were called the followers of Mahomet. Certainly, they subdued Arabia. The Saracens and Persians invaded Syria and Egypt, Chaldea, and Armenia. Afterward arose the Turks and Tartarians, receiving the religion of Mahomet, who have subdued in a manner all the provinces of the Roman empire in the east and to the south.

\index{Daniel (Book of)}\index{Rome}By the river Euphrates—famous throughout all of Asia—floated Babylon, the seat of the Eastern Monarchy. The mighty peoples of the East, the Assyrians, Babylonians, Medes, and Persians, who had ruled the world before the monarchies of Greece and Rome, seemed to be drowned, buried, and hidden, even bound within that same river. For the Macedonians of the West were governors of the world, and after them the Romans; and these mighty nations that we now name served them. But after the golden altar, as I said, was defiled, and countless people in the East and West revolted from the true Christian faith, God stirred up again the Eastern destroyers of the world, which had lain dormant as if asleep. For the prophets testify that these nations were the scourges of the world; therefore, God brought them forth again by his just judgment. Certainly, we read in the tenth chapter of Daniel that there was an Angel of Greece and an Angel of Persia, and that by them the entire people are understood. So now the nations of the East—Arabs, Saracens, Turks, and Tartarians—are raised up, who, for sin, might devastate the world, and the East might rule again, as Lactantius, from the Sibyl, prophesied would come to pass: let the West serve.

Let us learn from this discussion that all evils, and especially the devastation and ruin of kingdoms, arise from turning away from true religion to false religion. The foolish people of today judge the matter completely differently, and for that very reason are miserably destroyed. Let us learn that fierce nations are restrained and held back by God, so that they should not harm; that God stirs them up to inflict just punishment on the unrepentant. Thus it was with Sennacherib, Shalmaneser, and Nebuchadnezzar, who were called the servants of God and executed his judgments. Therefore, let us fear God and persevere in the true religion.

Moreover, the Saracens, Turks, and Tartarians are most diligently described. First, a wonderful expedition and celerity are commended in them, the principal virtue in wars. He says they are ever ready at every moment to execute the judgments of God. Therefore, he rehearses all parts of time, every hour of the day. And so there is no security from them; you can be in no surety. They are by and by in armor, and come unlooked for; they invade and speed their matters most luckily. He adds that through their most cruel and speedy armies, the third part of men in the world should be slain. Verily, Asia, Africa, and Europe have felt the most cruel slaughters and destructions of the Saracens, Turks, and Tartarians, ever since the time of Mahomet unto our days, about the space of nine hundred and twenty years. And also, the priests of Mahomet are very quick and diligent to allure men into their errors, and they want no lucky success.

\index{Apocalypse (Book of)}\index{Erasmus of Rotterdam}The number is also noted in an infinite manner, and the noble says he, of the army of horsemen, is twenty times ten thousand. And Mirias is the number of ten thousand, so that two myriads of myriads should make twenty times a thousand Myriades. And so the old translator has read or translated it: and Erasmus, twenty times ten thousand. The Dutch translation has, many thousand thousands. Laurence Valla in his annotations upon the New Testament does interpret, as has the Dutch translation, thousand thousands. But however it be, certain it is by the conference of other places, that a certain number is put for uncertain, that is to say for exceedingly great: and to be signified, that the horsemen of the Saracens, Turks, and Tartarians should be innumerable. For we read in the seventh of Daniel, thousand thousands served him, and ten thousand millions stood before him. And he speaks of angels (whom he signifies to be innumerable) and of their ministry. So also in the fifth of the Apocalypse, I heard, says he, the voice of many angels, and thousand thousands saying with a loud voice, etc. Certainly, the histories testify that the Saracens came out of Spain into France in number four hundred thousand. Paulus Aemilius in the second book of the acts of Frenchmen, recites that Charles Martel overcame three hundred three score and fifteen thousand Saracens. And Matthias a Michon in the first book, eighth chapter, of Sarmatia in Asia, Tamerlanes, says he, had an army of twelve hundred thousand. Moreover, it is plain that there were never in any age or memory greater armies of horsemen led out of any nation than of Turks, Saracens, and Tartarians. John adds that he heard their noble, either for that he would so confirm that he had said how their power should be greatest, or for that he would partly signify that their victories also were numbered, and should have an end. That in Daniel is most notable, Mene, Tekel Pheres: that is, has numbered, has weighed, has divided. He has numbered, says he, your kingdom, and has brought it to an end.

And at the beginning and also in the times that followed, the matters of Muhammad increased exceedingly. After Muhammad himself, they had in order twenty-five emirs [? rulers], for so they called their kings or princes, who ruled with great power until the year 860. About this time, the fifteenth emir, called Muhammad, sought to drive out and oppress Imbrael, the governor of Babylon, who sent for Muchulet the Turk out of Scythia against the emir. And the Turk dispatched his forces, and drove many of the Saracens out of Asia, and the Turks began to reign in the East. And the Saracens expelled from those regions came into Africa; from thence sailing into Sicily and other islands, they possessed Spain also, and overthrew nations never before conquered, and invaded Italy, spoiling Rome and consuming with fire many fine buildings. Concerning this matter, one may read Volateran in the twelfth book of Geography, in the threefold Arabia. About the year of our Lord 1300, the Turkish emperors began the Ottoman dynasty, who possess at this day a great part of Asia, Africa, and Europe. Baptista Ignatius has written about this in the end of the second book of the Roman emperors, and Paulus Iovius. Many of the Tartarians received the religion of Muhammad, and have most grievously afflicted the world, as Mathias a Michon writes in Sarmatia of Asia. And there is no doubt that the people of Muhammad have been of very great power, and are so still even at this day.

\index{Papacy}Now the horsemen and the horses on which they ride are also depicted, that is to say, the manners and power of the Mahometans are described. The horsemen had armor, not of iron, but fiery, yellow and brimstone-colored. Therefore, fire, hyacinth, and sulfur served as their breastplate, their armor. For the hyacinth immediately suggests smoke. The hyacinth, in color, resembles smoke next to fire and flame. And the horses had lion heads and serpent tails with heads. The horses breathed out of their mouths fire, smoke, and sulfur. With these plagues, he says, namely fire, smoke, and brimstone, a third part of men were slain. They also hurt me with the serpentine tails. He adds, their power was in their mouth, and the hurt in the tail. These appear to be understood and explained both spiritually and corporally. For the Mahometans, by their wicked doctrine, which is aptly compared to fire destroying, to smoke blinding, and to stinking sulfur, have destroyed countless people. Finally, with a lionish or tyrannical force, they have compelled many people to receive their Alcoran. And moreover, whenever their false prophets seem to flatter—for Isaiah says a false prophet is a tail—they act like serpents and infect people with the most corrupt poison of doctrine. Out of their mouths proceed not only blasphemous laws, but also marvelous praises, great boastings of victories, abominable blasphemies. Where do they say, is your Christian faith? Our religion of Mahomet overcomes all. All your things are miserable. Being vanquished, they serve like bound slaves everywhere. The thing itself declares that our religion is true and yours starkly false. Indeed, the Mahometans reign in a manner everywhere, in victories and riches they are fortunate and noble. That thing makes Christians afraid and causes many to revolt. For what is done among us is manifest to all. The gospel preachers have fought unluckily once or twice and endure great persecutions every hour. The papists overcome and rejoice. Therefore, there are many thousands who say, “How the thing itself speaks, whether religion is better.” Doubtless, this great felicity holds many still in error, who would otherwise be gentle and tractable. Therefore, it is no marvel that Turks or Mahometans prevail greatly with their mouth, since among Christians, victories and the felicity of this world are of such force in manner with all men. And yet they excel with their mouth and boastings, in the thing itself and in truth. For although the Turks are victorious, yet their religion is most false, most wicked, and most absurd.

And physically, how those things may be understood, there is no man who does not see, who knows Turkish histories. The Muslims burn with fire and brimstone, for scarcely is there any other nation which has so wasted the world with fire as this. Whichever way they turn, all things burn with a fierce fire, all is full of smoke. Their princes are lions, and their government is like lions, all tyrannical. They command cruel things, and nothing but bloody words come from their mouths. Therefore, many of them have called themselves the wrath of God and the whip or scourge of God. And truly, this wrath of the Lord follows corrupt doctrine and swerving from the faith. With these three plagues—fire, smoke, and brimstone—one-third of the world is slain and destroyed.

Moreover, the serpent’s tail admonishes chiefly that they do harm very much. For if the Muslims or Turks have entered into league with Christian princes anywhere, they have not done so without craft and guile. Those who have believed their promises and flattering words, and have sought and received aid from them, have nurtured a serpent in their bosom.

Hereof remain two notable examples. A discord arising between the emperor of Constantinople and his princes. Whilst Marcus, lord of Bulgaria, joined himself with the princes, or lords of Greece, the emperor was compelled to require aid of Amurathes the first of that name, the third Turkish emperor after Ottoman. And he aided the emperor gently. For he sent into Greece twelve thousand chosen Turks, with whom the emperor being aided, he discomfited and put to flight Marcus himself and the rest of the rebels. But that same amity was the beginning of the destruction of the empire of Constantinople, and of all the calamities of Greece. For when Amurathes understood by the soldiers which returned home that Greece was both a most goodly country, and not strong, by reason of the discord and dissension of princes, he determined to transport thither immediately, under pretense of persecuting the emperor’s enemies. And so began to possess Greece itself, which both his sons and nephews within a hundred years brought wholly into their subjection. In our time arose a discord for the realm of Hungary between Ferdinand, who now is emperor, and John Vayuode, prince of Hungary: which being not able in strength to match Ferdinand, was driven to crave aid of Solyman, emperor of Turks. The Turk was by and by ready with great faith and diligence placing John in his kingdom; howbeit we see that immediately he being extinct, the Turk enjoyed the kingdom of Hungary. Would God therefore that Christian princes would not trust the Turkish navy and warfare. For whilst the Mahometan laughs upon the Christian with a friendly countenance, he intends to put a serpent into his bosom, and to destroy him. And we are also at this day in this sixteenth, as also in the fifteenth fight, in the Papistical and Mahometical corruption, wickedness, and tyranny. The Lord Jesus deliver us from all these evils by his glorious coming unto judgment. Amen, Amen.
